The Neuron-specific IIS/FOXO Transcriptome in Aged Animals Reveals Regulatory Mechanisms of Cognitive Aging

Version 1

Abstract

Abstract

Cognitive decline is a significant health concern in our aging society. Here, we used the model organism C. elegans to investigate the impact of the IIS/FOXO pathway on age-related cognitive decline. The daf-2 Insulin/IGF-1 receptor mutant exhibits a significant extension of learning and memory span with age compared to wild-type worms, an effect that is dependent on the DAF-16 transcription factor. To identify possible mechanisms by which aging daf-2 mutants maintain learning and memory with age while wild-type worms lose neuronal function, we carried out neuron-specific transcriptomic analysis in aged animals. We observed downregulation of neuronal genes and upregulation of transcriptional regulation genes in aging wild-type neurons. By contrast, IIS/FOXO pathway mutants exhibit distinct neuronal transcriptomic alterations in response to cognitive aging, including upregulation of stress response genes and downregulation of specific insulin signaling genes. We tested the roles of significantly transcriptionally-changed genes in regulating cognitive functions, identifying novel regulators of learning and memory. In addition to other mechanistic insights, comparison of the aged vs young daf-2 neuronal transcriptome revealed that a new set of potentially neuroprotective genes is upregulated; instead of simply mimicking a young state, daf-2 may enhance neuronal resilience to accumulation of harm and take a more active approach to combat aging. These findings suggest a potential mechanism for regulating cognitive function with age and offer insights into novel therapeutic targets for age-related cognitive decline.

Introduction

The loss of cognitive function is a rising problem in our aging society. A 2008 study estimated that at least 22.2% (about 5.4 million) individuals over the age of 71 in the United States have at least mild cognitive impairment1–3. Furthermore, global dementia cases are predicted to triple from an estimated 57.4 million cases in 2019 to 152.8 million cases in 20504,5. As most industrialized countries are experiencing a rapid increase in the proportion of the aged population, understanding and potentially preventing the underlying issues of neuronal structural and behavioral decline associated with aging is crucial for societal health.

C. elegans is an excellent model system for studying neuronal aging, given its tractable genetics, short lifespan, and simple nervous system6. Most importantly, C. elegans experiences rapid loss of learning and memory with age7: by Day 4 of adulthood, all long-term associative memory ability is lost, and by Day 8, C. elegans cannot carry out associative learning or short-term associative memory7 – despite the fact that these worms can still move and chemotaxis perfectly well. That is, with age worms first lose long-term memory ability (by Day 4), then short-term memory and learning ability (Day 6-8), then chemotaxis (Day 10-12), then motility (Day 16)7,8. Because learning and memory decline extremely early, we consider worms that are only a week old to already be “cognitively aged,” despite the fact that they can chemotax and move well, and will continue to live for another one to two weeks. Therefore, we can examine neurons from these 7-8 day-old adults to explore the causes of these cognitive declines in animals that are otherwise quite healthy. Many human neuronal aging phenotypes and genes of interest for mammalian neuronal function are conserved in C. elegans9, making discoveries in C. elegans possibly applicable to humans.

The Insulin/IGF-1-like signaling (IIS)/FOXO pathway was first discovered to play a role in longevity in C. elegans. The lifespan of daf-2/Insulin/IGF-1 receptor mutants is twice that of wild-type animals10, and this lifespan extension requires the downstream Forkhead box O (FOXO) transcription factor DAF-1610. DAF-16/FOXO controls the expression of many genes that contribute to longevity, including stress response, proteostasis, autophagy, antimicrobial, and metabolic genes11. As a conserved regulator, the IIS/FOXO pathway also regulates longevity in Drosophila, mice, and humans12–15. In addition to regulating lifespan, the IIS pathway regulates neuronal function via the FOXO transcription factor. In particular, C. elegans IIS/daf-2 mutants display DAF-16-dependent improved learning, short-term memory, and long-term memory7. While both young and old daf-2 adult worms display increased learning and memory relative to wild-type, the duration of this extension is not known, and the mechanisms by which daf-2 mutants maintain neuronal function in older worms is not yet understood. Compared to wild-type worms, daf-2 mutants maintain maximum velocity8, motility16,17, neuromuscular junctions, the ability to regenerate axons 18,19, and neuronal morphology better with age20–22. (In particular, we previously showed that while daf-2 has lower observed motility on food23, this apparent is due to its high levels of a food receptor, ODR-108, and its downregulation reveals the much higher mobility of daf-2 animals8, even on food, in addition to its much higher and maintained maximum velocity with age.) Previously, we found that daf-2 worms also extend learning beyond wild-type’s ability7, but the full duration of this extension with age was not known. That is, exactly how late in life daf-2 mutants can still learn and remember, and whether this is proportional to their lifespan extension, was not previously determined.

We previously performed neuron-specific RNA-sequencing in young (Day 1) adult C. elegans and identified neuron-specific targets24; genes upregulated in daf-2 mutant neurons are distinct from those in the whole animal, and we found that these neuronal genes are necessary for the observed improvements in memory and axon regeneration in daf-2 mutant worms. However, whether daf-2 uses the same or different genes in young and old worms to improve and maintain cognitive function with age is unknown. To identify the transcriptional differences in the aging nervous system that might contribute to the loss of neuronal function with age in wild-type worms and the differences responsible for the extended abilities of daf-2 animals, here we RNA-sequenced FACS-isolated neurons of aged (Day 8) wild-type and IIS/FOXO mutants. (Some recent datasets using whole-animal single-cell RNA-seq have been generated for wild-type and daf-2 worms, and these are sufficient for whole-body aging and pseudobulk analyses25–27, but we have found that those data are not deep enough to use specifically for in-depth analysis of neurons, which can be difficult to gather from whole animals (Fig. S6e-h).

Other data are from larval stages and cannot be extrapolated to aging adults28.) To further investigate the role of the neuronal IIS/FOXO pathway, we identified genes both upregulated by the IIS/FOXO pathway, and genes that are differentially expressed in daf-2 mutants with age. We found that daf-2 differentially-regulated genes in the aged neurons are different from young neurons; in fact, many of these Day 8 daf-2 vs daf-16;daf-2 upregulated genes are stress response and proteolysis genes that may promote neuronal function and health. We then used functional assays to assess the contributions of daf-2-regulated genes to learning and memory. Our results suggest that daf-2’s neuronal targets in older worms are required to maintain neuronal functions with age, suggesting that additional and alternative mechanisms are at work in these aged mutants from their young counterparts.

Results

Wild-type neurons lose their neuronal function and identity with age

Previously, we found that cognitive abilities in C. elegans, including learning, short-term memory, and long-term memory, all decline with age7. Moreover, neuronal morphology and regeneration ability are also impaired with age18,20–22. However, how these phenotypes are regulated at the molecular level in aging neurons remains to be systematically characterized. Therefore, we were interested in first identifying gene expression changes with age in wild-type neurons to characterize the normal physiological aging process. Before choosing timepoints to assess neuronal transcriptome changes, we carried out associative learning and short-term associative memory assays7 as we have previously described7,29–31. Briefly, well-fed worms are starved for one hour, then re-fed while exposed to the neutral odorant butanone for one hour; a choice assay between butanone and control immediately after training tests associative learning, while a choice assay after one hour of recovery on food-only plates tests short-term associative memory7. Adult Day 1 worms are fully developed, young, and healthy, while wild-type Day 7-8 worms, although still in their mid-life, have completely lost their learning and short-term memory abilities already by Day 77 (Fig. 1a), thus we consider them “aged” for the purposes of understanding loss of cognitive ability. Therefore, we reasoned that a comparison of adult wild-type Day 1 neurons with wild-type neurons that are at least aged Day 7-8 should reveal changes with age that result in loss of cognitive function.

Identifying neuronal aging targets in WT worms using neuron-specific RNA- sequencing.

(a) Wild-type learning and 1hr memory results on Day 1 and Day 7. Learning and memory results are represented as learning index (LI). Details of the LI calculation are explained in the methods. Learning, N = 10, memory, N = 5. ****: p < 0.01. Student’s t-test. (b) PCA plot for Day 1 (orange) and Day 8 (blue) neuronal bulk RNA-seq samples. (c) Volcano plot comparing age-associated differentially-expressed genes in WT neurons. Genes downregulated with age (orange) and upregulated with age (blue) were obtained by neuron-specific RNA sequencing of adult wild-type animals with neuron-specific GFP expression. Adjusted p-value < 0.001, log2(Fold-change) > 2. N = 6 biological replicates per age. 1146 genes were significantly downregulated with age (higher in young neurons) and 2016 genes were upregulated with age (higher in old neurons) (d) Tissue prediction scores for genes higher in young neurons. (e) GO terms of genes that decline with age in wild-type neurons. Synaptic and signaling GO terms are enriched in neuronal genes. p-value calculated using hypergeometric distribution probability. (f) Comparison of whole-body higher-in-young genes and neuronal higher-in-young genes. GO Terms and representative genes were performed using g:Profiler software. P-value of overlapping regions were calculated using a hypergeometric calculator. (g) Normalized reads of ins-6, unc-4, mec-7, folt-2, fbf-1, and madd-4, in Day 1 and Day 8 neurons in our dataset. P- adjusted values were calculated from DESeq2 software.

To identify genes that regulate age-related morphological and functional decline in wild-type neurons, we performed neuron-specific transcriptomic analysis using our previous FACS neuronal isolation method24 on six biological replicates each of Day 1 and Day 8 adult wild-type worms (100,000 GFP+ cells were collected for each sample; Fig. S1a-c). Because we previously found that whole-worm analyses mask changes found specifically in neurons24, to complement our aging neuron studies, we also carried out RNA-sequencing analyses of aging whole worms (Fig. S2a-e), which we found is dominated by changes in the extracellular matrix (Fig. S2b), stress response/pathogen genes (Fig. S2c) and the alimentary system (intestine) (Fig. S2d), overshadowing neuronal changes.

Principal components analysis of the FACS-isolated neuron RNA-seq samples indicated that they are well separated by age (Fig. 1b), and downsampling analysis32 (Fig. S1e) suggested that we have sequenced to saturation, with an average of 41,636,463 uniquely-counted reads and detected the expression of 19725 coding and non-coding genes (log10(TPM)>0.5) (Fig. S1d). Enrichment analysis of genes that are differentially expressed with age (Fig. 1c-e) suggested that neuronal sorting and sequencing was successful (i.e., the fact that the sequenced genes are enriched for neuronal genes, as expected, and less enriched for all other major tissues indicates that our sorting is successfully selective for neurons). Tissue enrichment analysis of differentially-expressed genes suggested that aging neurons lose genes most expressed in the nervous system and neurons (Fig. 1d), as one might expect. Gene ontology (GO) analysis suggested that genes declining with age in neurons encode proteins important in neuronal function (Fig. 1e, f), including synaptic proteins (e.g., srh-59, rab-3, sng-1, sup-1), potassium channels (e.g., egl-23, twk-7, twk-49, ncs-5), and transmembrane transporters (e.g., folt-2, ccb-2, unc-79, exp-1). The decrease in expression of these genes during aging may indicate that neurons are losing their identity and their ability to perform neuronal functions, such as signal transduction and axonal transport, and correlates with the behavioral and morphological declines observed in aging wild-type worms.

Comparing whole-worm sequencing and neuron-specific sequencing (Fig. 1f), we found that genes involved in metabolic processes decline with age only in the body, and genes encoding structural proteins, lipid localization, and muscle system processes decline with age in both the body and in neurons, while neurons specifically lose genes that are associated with neuronal function, including synaptic proteins, neuropeptide signaling, and other neuron functions, correlating with neuronal loss of function with age. Together, these results indicate that neurons harbor many unique age-related changes that could be overshadowed in the whole-worm transcriptome but are revealed by neuron-specific sequencing.

Many genes that are more highly expressed in young neurons are known to be specific to a subset of neurons. ins-6, an insulin-like peptide specific to the ASI, ASJ, and AWA neurons28 that regulates longevity33 and aversive learning34, is significantly downregulated with age (Fig. 1g). srd-23, a serpentine receptor located at the AWB neuron cilia35, also decreases expression with age (Fig. S1f). Furthermore, various genes specific to sensory neurons (txt-12, flp-33), touch neurons (mec-7), and motor neurons (unc-4) decline in expression with age (Fig. 1g, Fig. S1f). Previous studies showed that loss of genes including ins-634, mec-736, unc-4, folt-219, madd-437, and fbf-131 lead to behavioral dysfunction in motility and chemosensory abilities; therefore, the decreased expression of these neuron type-specific genes with age may impact the function of individual neurons and disrupt neural circuit communication, ultimately contributing to the declines in behavior observed during aging (Fig. 1g).

As neurons age, genes that increase in expression, while assigned to the nervous system (Fig. 2a) are not specific for neuron function; instead, aged wild-type neurons express higher levels of many predicted F-box genes with predicted proteasome E3 activity (e.g., F-box proteins fbxa-158, fbxb-51, pes-2.1, and SKp1-related proteins skr-12, skr-6). Some transcription regulation (e.g., ced-13, tbx-43, nhr-221, end-1), and chromatin structure and function (e.g., his-54, dot-1.2, jmjd-3.2, hil-7, utx-1) genes also increase with age (Fig. 2b), even though neurons appear to lose their neuron-specific transcriptional identity with age.

Genes that increase with age cause behavioral defects.

(a) Tissue Query for wild-type genes expressed at higher levels in aged worms show lower nervous system and neuron prediction score. (b) GO terms of genes expressed higher in aged neurons highlight transcription regulation and proteolysis. GO term analysis was done using Wormcat 2.0. (c) Normalized reads of utx-1 on Day 1 and Day 8. (d) Short-term associative memory (STAM) assay shows that adult-only utx-1 knockdown improves 1hr and 2hr memory of wild-type worms on Day 2. (e) Normalized reads of ins-19 on Day 1 and Day 8. (f) ins-19 mutation improves learning and memory in STAM on Day 3 of adulthood. (g) Normalized reads of nmgp-1 on Day 1 and Day 8. (h) nmgp-1 RNAi knockdown improves memory in STAM on Day 2. P-adj value of normalized count change generated from DEseq2 analysis. N = 5 plates in each behavioral experiment. Representative result of 3 biological repeats is shown. *: p<0.05. **: p<0.01. ***: p<0.001. ****: p<0.0001

One ongoing discussion about changes during aging is how to interpret an increase in expression with age. There are two main models for genes that increase their expression with age and have a resulting impact on function: that they rise with age to compensate for lost function (“compensatory”) and therefore promote function, or that their expression is deleterious to function and only rises with age through dysregulation. If a gene is compensatory, then its knockdown would abrogate learning and memory, even in young animals. If a gene’s function is harmful to neurons, reducing its expression might be beneficial to the worm, even in young animals. To test this hypothesis, we reduced expression of a small set of highly-upregulated candidate genes in categories that might function in a compensatory manner. These include utx-1, a histone demethylase known to play a role in development38 and lifespan in worms39–41, and whose homolog has been implicated in cognition in mammals42,43; ins-19, an insulin-like peptide; and nmgp-1, a neuronal glycoprotein involved in chemosensation44. In each case, we see that gene expression is significantly higher in old than in young neurons (Fig. 2c, e, g). If a gene increases expression to benefit neurons, we would expect to see no difference in memory in young animals where there is no defect; by contrast, if the increase of a gene is deleterious, we would expect to see an improvement in behavior when knocked down, even in young animals. (When an RNAi strain is available, we use adult-only knockdown to prevent any possible deleterious effects caused by changes during development, which largely takes place in early larval stages; and testing in young adult animals is logical because there is no memory in aged wild-type, so any deleterious effect of knocking down a potentially compensatory gene in an aged would not result in a change. For all behavioral assays, we first prioritized significantly-changed genes with high fold-change, and then those with mammalian homologs.)

We found that 48 hrs (L4-Day 2) adult-only knockdown of utx-1 increases 1hr and 2hr memory (Fig. 2d), the loss-of-function mutation of ins-19 increases both learning and memory (Fig. 2f) and the adult-only knock-down of nmgp-1 extends memory at 2 hours (Fig. 2h). That is, in each of these cases, reduction of these genes did not impair memory, as loss of a compensatory function would appear; rather, loss of these age-upregulated genes improved wild-type memory. These results indicate that at least some neuronal genes that increase with age can have a negative impact on learning and memory, as demonstrated by the improvement of memory when knocked down, even in young animals. While it is still possible that some upregulated genes may act in a compensatory manner, the simplest model is that at least some are actively deleterious. We previously observed that loss of single genes that play a role in complex behaviors like learning and memory can have a large impact on function 60, unlike the additive roles of longevity-promoting genes 11. Therefore, one mechanism by which wild-type worms lose their function with age is not just by loss of neuronal gene expression, as one might expect, but also by dysregulation of expression of genes that can negatively impact learning and memory.

daf-2 mutants maintain learning and memory with age

We previously found that daf-2 animals have extended motility (Maximum Velocity) that correlates with and predicts their extension of lifespan8. Additionally, not only do young daf-2 worms have better memory than wild-type worms, but daf-2 mutants also maintain learning and memory better with age7,24. However, the duration of this improvement was unknown. To determine the proportion of life that worms can learn and remember, we tested wild-type, daf-2, and daf-16;daf-2 worms for their learning and associative memory ability every day until these functions were lost. We found that while wild-type worms lose their learning and short-term memory abilities by Day 7-8 (Fig. 1a, Fig. 3a, b), learning and memory span were significantly extended in daf-2 mutants (Fig. 3a, b); thus, a comparison of daf-2 neurons with wild-type neurons at Day 8 should reveal differences relevant to cognitive aging. The extension of learning and memory is dependent on the FOXO transcription factor DAF-16 (Fig. 3a); in fact, while daf-16;daf-2 mutants still have the ability to learn for a few days, these mutants are completely unable to carry out any memory ability, even on Day 1. Thus, learning ability, which is similar in wild-type and daf-2;daf-16 mutants, is mechanistically distinct from short-term memory ability31. daf-2 worms maintained learning ability until Day 19 and short-term (1hr) memory ability until Day 15, more than twice the duration of wild-type worms, while daf-16;daf-2 worms exhibit no short-term memory ability, even on Day 1 of adulthood (Fig. 3b). Our data suggest that the learning span-to-lifespan (Fig. S3c) and memory span-to-lifespan ratios in daf-2 worms were similar to or slightly higher than that of wild-type worms (Fig. 3c, d), indicating that daf-2 mutants maintain cognitive function for at least proportionally as long as wild-type worms do. Thus, daf-2 mutants maintain their higher cognitive quality of life longer than wild-type worms, while daf-16;daf-2 mutants spend their whole lives without memory ability (Fig. 3d), in contrast to claims that daf-2 mutants are less healthy than wild-type or daf-16 worms23. Additionally it should be noted that because our choice assays distinguish motility function from learning and memory function29, the improvements in memory with age shown by daf-2 mutants relative to wild type are distinct from daf-2’s improvements in motility that we previously showed8. Therefore, we are interested in these genes that might contribute to the extended cognitive function that daf-2 worms demonstrate.

Identifying neuronal IIS/FOXO targets in aged worms using neuron-specific RNA- sequencing.

(a) daf-2 mutants show better learning maintenance with age compared to N2 and daf-16;daf-2 worms. N = 10 plates in each condition. (b) daf-2 mutants show better memory maintenance with age compared to N2 worms. daf-16;daf-2 worms do not have 1hr memory on Day 1 of adulthood. N = 10 plates in each condition. (c-d) daf-2 mutants have a slightly larger learning span/lifespan ratio and memory span/lifespan ratio than N2 (wild type). Lifespan shown in Figure S3c. (e) PCA plot of Day 8 N2, daf-2, and daf-16;daf-2 neuronal RNA sequencing results. (f) Volcano plot of neuronal daf-2-regulated, daf-16-dependent up- and downregulated genes on adult Day 8 (Adjusted p-value < 0.05, log2(Fold-change) > 0.5, N = 6 biological replicates per strain). 570 genes were significantly upregulated and 814 genes were downregulated in daf-2 neurons compared with daf-16;daf-2. (g) Volcano plot of whole-worm daf-2 vs daf-16;daf-2 differentially-expressed genes during aging. 3154 genes are higher in daf-2, 1289 genes are higher in daf-16;daf-2 (log2[Fold-change(daf-2 vs daf-16;daf-2)] >1.5, p- adjusted <0.01). (h) Comparison of neuronal and whole-worm Day 8 daf-2 differentially-expressed genes (overlap p = 3.34E-63, hypergeometric test). Neuron-specific and shared daf-2 upregulated genes with the highest fold-changes are labeled.

Aging IIS/FOXO neurons express stress-resistance genes to maintain neuronal function with age

To identify genes that may improve memory and slow cognitive aging in long-lived daf-2 mutants, we compared the transcriptional profiles of Day 8 FACS-isolated neurons from daf-2 animals with Day 8 FACS-isolated wild-type and daf-16;daf-2 neurons; by Day 8, wild-type and daf-16;daf-2 worms have already lost their learning and memory ability, but daf-2 worms still maintain their cognitive functions (Fig. 3a, b). It should be noted that wild-type worms still have normal chemotaxis and motility at Day 87, and there is a separation of several days between the loss of cognitive functions and the loss of motility7; therefore, comparison of the neuronal transcriptomes of daf-2 with wild type and daf-16;daf-2 at this age should specifically highlight genes that are required for learning and memory rather than other functions.

The PCA of the daf-2, daf-16;daf-2, and wild-type neuronal Day 8 transcriptomes (Fig. 3e) indicates that aged daf-16;daf-2 mutant neurons are similar to aged wild-type neurons, correlating well with their similarly worsened cognitive functions at this age; that is, at a transcriptomic level, aged (Day 8) wild-type neurons and aged daf-16;daf-2 neurons are similar, which is echoed by their shared inability to carry out learning and memory by Day 8 of adulthood (Fig. 3ea, b). By contrast, the transcriptomes of aged daf-2 mutant neurons are distinct from both aged wild-type and aged daf-16;daf-2 neuron transcriptomes, just as the cognitive abilities of daf-2 are much greater than wild type or daf-16;daf-2 at this age. Downsampling analysis shows that our sequencing depth is sufficient to saturate the detectable differential expression (Fig. S3g, h). We obtained an average of 47,233,119 counted reads per sample (Supplemental Table 5) and detected the expression of 16,488 coding and non-coding genes (Fig. S3e).

We identified 570 upregulated and 814 downregulated genes in Day 8 daf-2 neurons compared to Day 8 daf-16;daf-2 neurons (Fig. 3f). A large fraction of the downregulated genes in Day 8 daf-2 vs daf-16;daf-2 neurons are “nematode-specific peptide family” (nspc-) genes of unknown function (Fig. 3f). While the daf-2 vs daf-16;daf-2 changes in whole worms largely replicated the results from our previous studies of young animals11,45 (Fig. 3g, Fig S6a), comparison of the daf-2 vs daf-16;daf-2 differential transcriptional changes in Day 8 whole worms and Day 8 neurons reveal shared (174 genes) and neuron-specific gene expression changes (396 genes; dod-24, srh-2, lin-42, etc.) (Fig. 3h). Not surprisingly, previously-identified genes from whole-worm daf-2 vs daf-16;daf-2 and N2 (e.g., sod-3, mtl-1, cpi-1, hsp-12.6, etc.) that play roles in both neurons and other tissues even in Day 1 daf-2 mutants appear in the shared list. (Some neuron-specific Day 8 daf-2-upregulated genes have not been reported to be expressed in neurons previously (e.g., spin-2), further suggesting the value of transcriptomic analyses of isolated neurons in mutant backgrounds at this age.)

Many genes upregulated in Day 8 daf-2 neurons relative to daf-16;daf-2 are related to stress responses, including heat stress (e.g., hsp-12.6, hsp-12.3, F08H9.4/hsp), oxidative stress (e.g., sod-3), and metal stress genes (e.g., mtl-1); and proteolysis (e.g., cpi-1, cpr-2, and tep-1). The upregulation of these genes may perform neuroprotective functions, as their homologs in mammals have been shown to do (Table 1). Specifically, 36 of the top 100 upregulated genes have identified orthologs or identified domains with known function, of which 32 of (89%) have functions in promoting neuronal health. These mammalian homologs protect neurons against protein aggregation and harmful metabolites (e.g., cpi-1, alh-2, ttr-41, gpx-5)46–50, maintain synaptic organization and neuronal homeostasis (e.g., dod-24, ptr-19, plep-1)51–53, facilitate neuronal injury repair (e.g., F08H9.4, sod-3)54,55, and maintain normal neuronal function (e.g., lgc-28, slc-36.3, lin-42)56–59. Together, these genes may help maintain daf-2’s neuronal health and protect neurons from accumulation of environmental harm during aging.

IIS/FOXO transcriptomic changes are necessary for daf-2 mutant’s improved neuronal functions

We were interested not only in the genes that remained upregulated with age, but also in genes that might have increased with age in the high-performing daf-2 mutants. That is, are there genes that increase in expression in daf-2 mutants that are necessary or beneficial for their continued high performance with age? Some of the Day 8 daf-2 vs wild-type or daf-16;daf-2 upregulated genes are also Class 1 DAF-16-dependent genes11 (sod-3, hsp-12.3, fat-5, and mtl-1, hil-1, and dao-2). However, many more genes were differentially expressed in Day 8 daf-2 vs daf-16;daf-2 neurons from our Day 1 data24 (Fig. 4a). Of the “new” genes – that is, genes upregulated specifically in neurons of Day 8 vs Day 1 of daf-2 vs daf-16;daf-2 – many have mammalian homologs that have been shown to play neuroprotective roles, by protecting against aggregation proteins and harmful metabolites, maintaining synaptic organization, neuronal homoeostasis, or neuronal activity, or facilitating neuronal injury repair (see Table 1 for specific references).

Neuronal IIX/FOXO aging targets regulate memory decline with age in daf-2 worms.

(a) Comparison of neuronal Day 1 and Day 8 daf-2 vs daf-16;daf-2 upregulated genes. All shared genes and top Day 8-specific daf-2 upregulated genes are labeled. (b) daf-2-regulated fold-change profile of candidate genes. All candidates are upregulated in daf-2 mutants. (c) Description of candidate genes. log2(Fold-change) and p-adjusted values from the daf-2 vs daf-16;daf-2 comparison unless stated otherwise. (d) Candidate gene knockdown effects on Day 6 adult daf-2 learning (0hr after conditioning). Two candidate genes, dod-24 and F08H9.4, show a significant decrease in learning ability. N = 5 plates in each condition, the representative image of 3 biological repeats shown. (e) Candidate gene knockdown effects on Day 6 adult daf-2 short-term memory (1hr after conditioning). C44B7.5, dod-24, F08H9.4, mtl-1, and alh-2 showed significant decreases in memory. N = 5 plates in each condition, the representative image of 3 biological repeats shown. *: p < 0.05. **: p < 0.01. ****: p <0.0001.

If the upregulated genes in aged daf-2 neurons are responsible for the extended memory span of daf-2 mutants, knocking down those genes should block older daf-2 mutants’ memory functions. Therefore, we tested the effect of RNAi knockdown of the top fold-change candidate genes on daf-2’s memory in aged adults. (We chose Day 6 for testing because by then, like on Day 8, wild-type worms have already lost their learning and most memory abilities, but daf-2 worms retain normal cognitive functions, but avoids the increased naïve chemotaxis that we observe in older daf-2 animals.) As shown in Fig. 4b-c, we selected these significantly differentially-expressed candidate genes based on their ranking in fold-change. Previously, we have found that the top significantly differentially-expressed genes (by fold-change) are most likely to have strong effects on function, while less differentially-changed genes have less of an effect11,24,60, therefore we prioritized genes that are significantly different and the most highly expressed in daf-2 mutants compared to daf-16;daf-2 mutants for subsequent testing (Fig. 4c). daf-2 worms, including neurons, are susceptible to RNA interference24,61. Of the eight candidate genes we tested, the reduction of two of them (F08H9.4 and dod-24, originally classified as a Class II gene with proposed immune activity) significantly decreased daf-2’s learning ability on Day 6 (Fig. 4d). Those genes plus reduction of three additional genes (C44B7.5, mtl-1, and alh-2) affected 1hr memory (Fig. 4e) in Day 6 daf-2 mutants. That is, knockdown of the heat shock-related gene F08H9.4, the innate immunity gene dod-24, aldehyde dehydrogenase alh-2 and previously uncharacterized gene C44B7.5 are required to some degree for daf-2’s extended memory ability, although only C44B7.5 had a major effect on memory. (The reduction of the metal stress gene mtl-1, which is expressed in neurons as well as the rest of the body, had no effect on learning and a slight effect on memory.)

One caveat of these experiments is that, while we found these genes through the isolation of neurons from aged worms and subsequent RNA-sequencing, the knockdown of the genes and its effects are not necessarily neuron-autonomous; however, alh-2 and F08H9.4 were reported to only be expressed in neurons and the cephalic sheath cell62, and C44B7.5 and dod-24, while expressed more broadly, were not upregulated in daf-2 in the whole-worm analysis (Fig. 3f), therefore their effects are most likely neuron-autonomous. In fact, dod-24 is one of the original Class 2 daf-2-downregulated genes from whole-animal analyses, suggesting that dod-24’s increase in expression is specifically in neurons, therefore the effect of its knockdown is most likely to be neuron-autonomous.

Together, these data suggest that the specific genes that are differentially regulated in Day 8 daf-2 mutants may aid in slowing neuronal function decline and behavioral changes associated with aging. Furthermore, memory maintenance with age might require additional functions that were not previously uncovered in analyses of young animals.

Discussion

Although it has been shown previously that daf-2 worms maintain various functions with age, how long they can maintain learning and memory with age, and the genes that might be responsible for these extended neuronal functions, have not been previously explored. Here we have found that daf-2 worms maintain learning and memory abilities proportional with (or even slightly beyond) their degree of lifespan extension, underscoring daf-2’s improved healthspan8. To understand how memory is lost with age and retained in insulin/IGF-1-like signaling mutants, we have characterized the neuronal transcriptomes of aged wild-type worms and IIS (daf-2) and IIS/FOXO (daf-16;daf-2) mutants (Fig. 5). We found that wild-type neuronal aging is characterized by a downregulation of neuronal function genes and an upregulation of proteolysis genes and transcriptional and epigenetic regulators, which together may help explain the loss of neuronal identity and function with age. We also identified the transcriptomic profile accompanying daf-2’s extended learning and memory span. Specifically, daf-2 neurons maintain higher expression of stress response genes and predicted neuronal homeostasis functions (Table 1), which may help make them more resistant to environmental adversities and age-related decline. We also identified genes responsible for wild-type worms’ worsened learning and memory with age.

Aged daf-2 neurons upregulate neuroprotective genes to maintain improved cognitive behaviors.

During normal neuronal aging, neuron-specific genes decrease in expression, while proteolysis and epigenetic regulators are upregulated, resulting in neuron dysfunction and cognitive function loss. In aged daf-2 mutants, upregulation of neuroprotective genes including dod-24, F08H9.4, C44B7.5, alh-2, and mtl-1 contribute to daf-2’s improved cognitive function. The diagram was generated using Biorender.

By employing a FACS-based neuron-sorting technique, we selectively analyzed neuron-function-related genes and investigated their aging process, which is not easily discernible through whole-worm sequencing. Sequencing many biological repeats of aging neurons to high depth with ribosomal RNA depletion allowed us to detect a larger number of genes compared to other neuron-related bulk and single-cell sequencing profiles26, providing a deep transcriptomic dataset of aged wild-type, IIS mutant, and IIS/FOXO mutant neurons. Our analysis allowed us to identify differentially-expressed genes that are known to be expressed in at a small number of neurons, even for low-abundance genes. Notably, our sequencing results uncovered genes previously not known to be expressed in neurons that remained undetected in other datasets. Moreover, we revealed the involvement of known neuronal genes in the aging process, such as ins-6 and srd-23. We hope that this dataset will become a valuable resource for detecting new candidates in neuronal aging.

For example, dod-24, which we observed to be upregulated in daf-2 neurons and required for daf-2’s extended memory, was downregulated in the daf-2 whole-worm transcriptome (Fig. S4f). dod-24 has been traditionally classified as a Class II gene that is downregulated in daf-2 worms and upregulated by daf-16 RNAi treatment11,45. Functionally, it has been shown to be an innate immunity gene upregulated during pathogen infection63–65, and its whole-body reduction has been shown to extend the lifespan of wild-type animals11. However, here we find that dod-24 is beneficial in the nervous system and required for daf-2’s extended learning and memory in aged worms. This intriguing contrast between the whole-worm transcriptome and the neuron-specific transcriptome suggests that some genes may have distinct regulatory roles in the nervous system, necessitating a more precise approach beyond whole-worm transcriptomics.

Using this neuron-specific sequencing profile of aged cells, we identified key pathways that change during neuron aging. Our sequencing of aged neurons uncovered active transcriptomic alterations during aging, resulting in not just transcriptional silencing but also upregulation of various pathways. We found that the Day 8 daf-2 vs daf-16;daf-2 neuronal differentially-expressed genes that we have newly discovered here differ from the neuronal Day 1 daf-2 vs daf-16;daf-2 dataset we previously obtained24. These Day 8 differentially-expressed genes are not canonical neuronal genes, such as receptors and ion channels; instead, there are more metabolic and proteolytic genes whose protein orthologs have been shown to be neuroprotective. All top 50 genes and 90% of the top 100 genes with identified mammalian orthologs have been shown to be essential to neuronal functions in mammals (Table 1). These results indicate that instead of simply mimicking a young state, daf-2 may enhance neuron’s resilience to accumulation of harm and take a more active approach to combat aging. These changes suggest that daf-2’s extended memory maintenance may require different mechanisms than function in young animals; daf-2 may maintain neuronal function not just by retaining a youthful transcriptome, but also by increasing the expression of genes that promote resilience, such as stress-response genes and proteolysis inhibitors.

In addition to examining aging in wild-type and IIS/FOXO mutants independently, our results further linked the normal aging process to altered gene regulation in the IIS pathway. utx-1, nmgp-1, and ins-19 increase in expression in aged neurons, and we found that their reduction improved memory, indicating that at least some of the genes whose expression rises with age can have a negative impact on normal cognitive functions, rather than acting in a compensatory manner. utx-1 is an H3K27me3 histone demethylase we found to be higher in wild-type aged neurons, but it is also involved in the IIS pathway. The downregulation of utx-1 has been shown to regulate development38 and promote longevity39–41, and its mammalian homolog has been implicated in regulating cognitive abilities42,43. The longevity response of utx-1 depends on daf-1639–41. The loss of utx-1 decreases methylation on the daf-2 gene, thus increasing DAF-16’s nuclear localization, mimicking a daf-2 mutation39. This example of the crosstalk between normal aging and IIS/FOXO mutants offers valuable insights into modifying the aging process for enhanced longevity and cognitive health.

We found that the insulin-like peptide ins-19 was upregulated in aged neurons and was downregulated in aged daf-2 neurons, and its downregulation in wild-type worms extended memory span. Insulin-like peptides play crucial roles as receptor ligands (in both agonist and antagonist roles) for DAF-2, and we have found them to be downregulated in daf-2 mutants compared with daf-16;daf-2 mutants, possibly creating a feedback loop that dampens the insulin signaling pathway, as was previously shown for ins-7 and ins-1811,66. These peptides exhibit diverse functions in development, dauer formation, and longevity11,66–70. Notably, certain insulin-like peptides have been linked to neuronal activities, such as the regulation of aversive learning by the two antagonistic peptides ins-6 and ins-734, and reduced long-term learning and memory by ins-22 RNAi60. In our study, the expression changes of ins-19 during wild-type aging and in daf-2 mutants provide an example of how longevity mutants can reverse wild-type transcriptional changes during aging, ultimately reducing behavioral and functional decline.

Summary of mechanistic insights

Our analysis of transcriptomes from isolated aged Day 8 neurons of wild-type, daf-2, and daf-16;daf-16 mutants reveal several major mechanistic insights. Specifically, we found that wild-type neurons lose their neuronal identity through a combination of the loss of neuron-specific function genes with age, and the concomitant dysregulated increase in non-neuronal genes with age. Furthermore, at least a fraction of the top-upregulated genes with age can play deleterious roles; that is, they rise with age, and their knockdown improves function, even in young animals. This argues against the idea that all of these genes are playing a compensatory role with age.

We also found that the knockdown of individual top-ranked genes that function in learning and memory can have a large impact - like removing a cog of a machine. This is in contrast to our earlier findings regarding gene reduction in lifespan, where most cellular longevity processes regulated by DAF-16 activity appear to be additive, and therefore loss of individual major genes downstream of DAF-2 and DAF-16 have at most a 5-10% impact11. Several of these genes we found to be required for daf-2’s age-related improvement in learning and memory - namely dod-24, F08H9.4, C44B7.5, and alh-2 – were previously not associated with memory function. Finally, these genes are distinct from the set of upregulated Day 1 daf-2 vs daf-16;daf-2 genes; how they each individually maintain neuronal function better with age will be interesting to dissect.

Conclusions

Beyond our sequencing analysis, we have established links between genomics, function, and behavior. We also identified several new genes required for daf-2’s age-related improvement in learning and memory, shedding light on their neuron-specific roles. These additional findings further suggest that neuronal sequencing datasets can be used to identify functional candidates and pathways during the aging process. By bridging the gap between transcriptomic landscapes, genetic regulation, and functional outcomes, our study provides a greater understanding of the mechanisms underlying neuronal aging, providing insights into development of aging interventions.

Methods

Strains and worm cultivation

N2 (wild type), OH441: otIs45(unc-119::GFP), CQ295: otIs45(unc-119::GFP);daf-2(e1370), CQ296: otIs45(unc-119::GFP);daf-16(mu86);daf-2(e1370), LC108: uIs69 (myo-2p::mCherry + unc-119p::sid-1), CQ705: daf-2(e1370) III, 3X outcrossed, CQ745: daf-2(e1370) III; vIs69 [pCFJ90(Pmyo-2::mCherry + Punc-119::sid-1)] V, QL188: ins-19(tm5155) II, CX3695: kyIs140(str-2::GFP + lin-15(+)), CQ461: (daf-2(e1370);Pmec-4::mCherry) and CQ501: (daf-2 (e1370);zip-5(gk646);Pmec-4::mCherry). Strains were grown on high-growth media (HGM) plates seeded with E. coli OP50 bacteria using standard methods 71.

Tissue-specific isolation

For neuronal isolation, five plates of fully-grown worms from HG plates were synchronized by hypochlorite treatment, eggs spread on seeded plates to hatch, and at least five plates/replicate were grown to L4 on HGM plates until transferred to HGM plates with FUdR to avoid progeny contamination. This gives us ∼6000 healthy Day 8 worms to sort. Neuron isolation and Fluorescent-activated cell sorting were carried out as previously described 24,62. Briefly, worms were treated with 1000uL lysis buffer (200 mM DTT, 0.25% SDS, 20 mM HEPES pH 8.0, 3% sucrose) for 6.5 mins to break the cuticle. Then worms were washed and resuspended in 500uL 20mg/mL pronase from Streptomyces griseus (Sigma-Aldrich). Worms were incubated at room temperature with mechanical disruption by pipetting until no whole-worm bodies were seen, and then ice-cold osmolarity adjusted L-15 buffer(Gibco) with 2% Fetal Bovine Serum (Gibco) were added to stop the reaction. Prior to sorting, cell suspensions were filtered using a 5um filter and sorted using a FACSVantage SE w/ DiVa (BD Biosciences; 488nm excitation, 530/30nm bandpass filter for GFP detection). Sorting gates were determined by comparing with age-matched, genotype-matched non-fluorescent cell suspension samples. Fluorescent neuron cells were directly sorted into Trizol LS. 100,000 GFP+ cells were collected for each sample.

RNA extraction, library generation and sequencing

We used standard trizol-chloroform-isopropanol method to extract RNA, then performed RNA cleanup using RNeasy MinElute Cleanup Kit (Qiagen). RNA quality was assessed using the Agilent Bioanalyzer RNA Pico chip, and bioanalyzer RIN>6.0 samples were observed before library generation. 2ng of RNA was used for library generation using Ovation SoLo RNA-Seq library preparation kit with AnyDeplete Probe Mix-C. elegans (Tecan Genomics) according to the manufacturer’s instructions 72. Library quality and concentration was assessed using Agilent Bioanalyzer DNA 12000 chip. Samples were multiplexed and sequencing were performed using NovaSeq S1 100nt Flowcell v1.5 (Illumina).

Data processing

FastQC was performed on each sample for quality control analysis. RNA STAR package was used for mapping paired-end reads to the C. elegans genome ce11 (UCSC Feb 2013) using the gene model ws245genes.gtf. Length of the genomic sequence around annotated junctions is chosen as read length ™1. 50-70% reads were uniquely mapped. Reads uniquely mapped to the genome were then counted using htseq-count (mode = union). DESeq2 analysis was then used for read normalization and differential expression analysis on counted reads 73. Genes with a log10TPM >0.5 were considered as detected and genes with a log2(fold-change) > 0.5 and p- adjusted <0.05 are considered differentially expressed in further analysis. Gene ontology analysis were performed using gprofiler74 or WormCat 2.075 and category 2 was selected to show. Tissue query was performed on top 500 highest fold-change genes, using worm tissue query website (www.worm.princeton.edu)62, and only major systems were selected in the analysis.

Learning and Memory Experiments

We performed Short-Term Associative Memory (STAM) experiments as previously described 7,29. Briefly, we used five plates of synchronized adult worms/sample to perform each experiment. One plate of worms was used to test naïve chemotaxis assay without conditioning, while the other 3 plates were washed into M9, and washed 3 additional times to get rid of the bacteria. These worms are starved for 1hr to prime them for food uptake. Then these worms are transferred to conditioning plates with NGM plates seeded with OP50 and 10% butanone stripes on the lid for 1hr to perform conditioning. After conditioning, worms are either transferred from the conditioning plate directly to the chemotaxis plates to assess learning, or transferred to holding plate for 1hr or 2hrs to assess for memory. After staying on holding plates for 1hr or 2hrs, worms are then transferred onto chemotaxis plates to assess for short-term memory. Chemotaxis assays were performed by transferring worms onto chemotaxis plates with 1uL 10% butanone and 1uL ethanol spots separated by 8cm on a 10cm NGM plate. Worms who have reached either butanone spot or the ethanol spot are paralyzed by the 1uL 7.5% NaN3 on these spots. For each timepoint, 5 chemotaxis plates are used to minimize the variation of the outcome. We performed this chemotaxis assay to butanone on naïve and appetitive-trained worms at different time points to assess change in preference to butanone. Chemotaxis index is calculated as (# of worms at butanone-# of worms at ethanol)/(total # of worms - # of worms at origin).

Learning index is calculated by subtracting trained chemotaxis index with naïve chemotaxis index.

For learning and memory span assays, we obtained synchronized worms from hypochlorite-treated eggs. Synchronized worms were washed onto 5’-fluorodeoxyuridine (FUdR) at L4 and maintained on FUdR plates by transferring to new plates every 2 days. 1 Day Prior to experiments, worms are washed onto fresh HG plates without FUdR to avoid change in behavior caused by FUdR. To verify that FUdR has no effect on short-term memory, we compared worms with and without FUdR (Fig. S6d), and found no differences. For utx-1 and nmgp-1 RNAi experiments, synchronized L4 wild-type worms were washed onto HGM plates with carbenicillin and IPTG and seeded with HT115 RNAi bacteria containing the RNAi constructs from the Ahringer Library. For daf-2 upregulated candidates’ RNAi experiments, synchronized L4 daf-2 RNAi-sensitive worms were washed onto HGM plates added with carbenicillin, FUdR, and isopropyl-b-D-thiogalactopyranoside (IPTG) and seeded with HT115 bacteria containing RNAi constructs generated from the Ahringer RNAi Library, then were transferred onto fresh RNAi plates every 2 days until Day 6. 1 Day Prior to experiments, worms are transferred onto plates without FUdR.

Quantitative and Statistical Analysis

All experimental analysis was performed using Prism 8 software. Two-way ANOVA with Tukey post-hoc tests were used to compare the learning curve between control and experimental groups. One-way ANOVA followed by Dunnet post-hoc tests for multiple comparisons was performed to compare learning or 2hr memory between various treatment groups and control. Chi-square test was performed to compare the neuron morphology change between young and aged AWC neurons. All GO term analysis were perform using Wormcat 2.0 software with Bonferroni corrected adjusted p-values. Venn diagram overlaps were compared using the hypergeometric test. Differential expression analysis of RNA-seq were performed using DESeq2 algorithm and adjusted p-values were generated with Wald test using Benjamini and Hochberg method (BH-adjusted p values). Additional statistical details of experiments, including sample size (with n representing the number of chemotaxis assays performed for behavior, RNA collections for RNA-seq, and the number of worms for microscopy), can be found in the methods and figure legends. Regression analyses were performed using sklearn packages. Correlations were calculated using the SciPy packages.

Data availability

Sequencing reads are deposited at NCBI BioProject under accession number PRJNA999305.

Code availability

Analysis codes generated from this study are available upon request.

Resource availability

Further information and requests for resources and reagents should be directed to and will be fulfilled by Coleen T. Murphy (ctmurphy@princeton.edu).

Supporting information

eLife Assessment

This important study investigates the transcriptional changes in neurons that underlie loss of learning and memory with age in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants. The presented evidence is convincing, utilizing a cutting-edge method to isolate neurons from worms for genomics that is clearly conveyed with a rigorous experimental approach. Overall, this study supports that older daf-2 worms maintain cognitive function via mechanisms that are unique from younger wild type worms, which will be of interest to neuroscientists and researchers studying ageing.

Reviewer #1 (Public Review):

The authors perform RNA-seq on FACS-isolated neurons from adult worms at days 1 and 8 of adulthood to profile the gene expression changes that occur with cognitive decline. Supporting data are included indicating that by day 7 of adulthood, learning and memory are reduced, indicating that this time point or after represents cognitively aged worms. Neuronal identity genes are reduced in expression within cognitively aged worms, whereas genes involved in proteostasis, transcription/chromatin, and stress response are elevated. A number of specific examples are provided, representing markers of specific neuronal subtypes, and correlating expression changes to the erosion of particular functions (e.g. motor neurons, chemosensory neurons, aversive learning neurons, etc).

To investigate whether the upregulation of genes in neurons with age is compensatory or deleterious, the authors reduced the expression of a set of three significantly upregulated genes and performed behavioral assays in young adults. In each case, reduction of expression improved memory, consistent with a model in which age-associated increases impair neuronal function. This claim would be bolstered by an experiment elevating the expression of these genes in young neurons, which should reduce the learning index if the hypothesis is correct.

The authors then characterize learning and memory in wild-type, daf-2, and daf-2/daf-16 worms with age and find that daf-2 worms have an extended ability to learn for approximately 10 days longer than wild types. This was daf-16 dependent. Memory was extended in daf-2 as well, and strikingly, daf-2;daf-16 had no short-term memory even at day 1. Transcriptomic analysis of FACS-sorted neurons was performed on the three groups at day 8. The authors focus their analysis on daf-2 vs. daf-2;daf-16 and present evidence that daf-2 neurons express a stress-resistance gene program. One question that remains unanswered is how well the N2 and daf-2;daf-16 correlate overall, and are there differences? This may be informative as wild type and daf-2;daf-16 mutants are not phenotypically identical when it comes to memory, and there may be differences that can be detected despite the overlap in the PCA. This analysis could reveal the daf-16 targets involved in memory.

The authors tested eight candidate genes that were more highly expressed in daf-2 neurons vs. daf-2;daf-16 and showed that reduction of 2 and 5 of these genes impaired learning and memory, respectively, in daf-2 worms. This finding implicates specific neuronal transcriptional targets of IIS in maintaining cognitive ability in daf-2 with age, which, importantly, are distinct from those in young wild type worms.

Reviewer #2 (Public Review):

Weng et al. perform a comprehensive study of gene expression changes in young and old animals, in wild-type and daf-2 insulin receptor mutants, in the whole animal, and specifically in the nervous system. Using this data, they identify gene families that are correlated with neuronal ageing, as well as a distinct set of genes that are upregulated in neurons of aged daf-2 mutants. This is particularly interesting as daf-2 mutants show both extended lifespans and healthier neurons in aged animals, reflected by better learning/memory in older animals compared with wild-type controls. Indeed, the knockdown of several of these upregulated genes resulted in poorer learning and memory. In addition, the authors showed that several genes upregulated during ageing in wild-type neurons also contribute to learning and memory; specifically knockdown of these genes in young animals resulted in improved memory. This indicates that (at least in this small number of cases), genes that show increased transcript levels with age in the nervous system somehow suppress memory, potentially by having damaging effects on neuronal health.

Finally, from a resource perspective, the neuronal transcriptome provided here will be very useful for C. elegans researchers as it adds to other existing datasets by providing the transcriptome of older animals (animals at day 8 of adulthood) and demonstrating the benefits of performing tissue-specific RNAseq instead of whole-animal sequencing.

The work presented here is of high quality and the authors present convincing evidence supporting their conclusions. I only have a few comments/suggestions:

(1) Do the genes identified to decrease learning/memory capacity in daf-2 animals (Figure 4d/e) also impact neuronal health? daf-2 mutant worms show delayed onset of age-related changes to neuron structure (Tank et al., 2011, J Neurosci). Does knockdown of the genes shown to affect learning also affect neuron structure during ageing, potentially one mechanism through which they modulate learning/memory?

(2) The learning and memory assay data presented in this study uses the butanone olfactory learning paradigm, which is well established by the same group. Have the authors tried other learning assays when testing for learning/memory changes after the knockdown of candidate genes? Depending on the expression pattern of these genes, they may have more or less of an effect on olfactory learning versus for example gustatory or mechanosensory-based learning.

(3) I have a comment on the 'compensatory vs dysregulatory' model as stated by the authors on page 7. I understand that this model presents the two main options, but perhaps this is slightly too simplistic: the gene expression that rises during ageing may be detrimental for memory (= dysregulatory), but at the same time may also be beneficial for other physiological roles in other tissues (=compensatory).

Reviewer #3 (Public Review):

Summary:

In this manuscript, Weng et al. detect a neuron-specific transcriptome that regulates aging. The authors first profile neuron-specific responses during aging at a time point where a loss in memory function is present. They discover signatures unique to neurons which validate their pipeline and reveal the loss of neuron identity with age. For example, old neurons reduce the expression of genes related to synaptic function and neuropeptide signaling and increase the expression of chromatin regulators, insulin peptides, and glycoproteins. The authors discover the detrimental effect of selected upregulated genes (utx-1, ins-19, and nmgp-1) by knocking them down in the whole body and detecting improvement of short memory functions. They then use their pipeline to test neuronal profiles of long-lived insulin/IGF mutants. They discover that genes related to stress response pathways are upregulated upon longevity (e.g. dod-24, F08H9.4) and that they are required for improved neuron function in long-lived individuals.

Strengths:

Overall, the manuscript is well-written, and the experiments are well-described. The authors take great care to explain their reasoning for performing experiments in a specific way and guide the reader through the interpretation of the results, which makes this manuscript an enjoyable and interesting read. Using neuron-specific transcriptomic analysis in aged animals the authors discover novel regulators of learning and memory, which underlines the importance of cell-specific deep sequencing. The time points of the transcriptomic profiling are elegantly chosen, as they coincide with the loss of memory and can be used to specifically reveal gene expression profiles related to neuron function. The authors showcase on the dod-24 example how powerful this approach is. In long-lived insulin/IGF-1 receptor mutants body-wide dod-24 expression differs from neuron-specific profiles. Importantly, the depletion of dod-24 has an opposing effect on lifespan and learning memory. The dataset will provide a useful resource for the C. elegans and aging community.

Weaknesses:

While this study nicely describes the neuron-specific profiles, the authors do not test the relevance in a tissue-specific way. It remains unclear if modifying the responses only in neurons has implications for either memory or potentially for lifespan. The authors point to this in the text and refer to tissue-specific datasets. However, it is possible that the tissue-specific profile changes with age. The authors should consider mining publicly available cell-specific aging datasets and performing neuron-specific RNAi to test the functional relevance of the neuron-specific response. This would strengthen the importance of cell-specific profiling.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment

This important study investigates the transcriptional changes in neurons that underlie loss of learning and memory with age in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants. The presented evidence is convincing, utilizing a cutting-edge method to isolate neurons from worms for genomics that is clearly conveyed with a rigorous experimental approach. Overall, this study supports that older daf-2 worms maintain cognitive function via mechanisms that are unique from younger wild type worms, which will be of interest to neuroscientists and researchers studying ageing.

Thank you, we appreciate the positive comments.

Public Reviews:

Reviewer #1 (Public Review):

The authors perform RNA-seq on FACS-isolated neurons from adult worms at days 1 and 8 of adulthood to profile the gene expression changes that occur with cognitive decline. Supporting data are included indicating that by day 7 of adulthood, learning and memory are reduced, indicating that this time point or after represents cognitively aged worms. Neuronal identity genes are reduced in expression within cognitively aged worms, whereas genes involved in proteostasis, transcription/chromatin, and stress response are elevated. A number of specific examples are provided, representing markers of specific neuronal subtypes, and correlating expression changes to the erosion of particular functions (e.g. motor neurons, chemosensory neurons, aversive learning neurons, etc).

To investigate whether the upregulation of genes in neurons with age is compensatory or deleterious, the authors reduced the expression of a set of three significantly upregulated genes and performed behavioral assays in young adults. In each case, reduction of expression improved memory, consistent with a model in which age-associated increases impair neuronal function. This claim would be bolstered by an experiment elevating the expression of these genes in young neurons, which should reduce the learning index if the hypothesis is correct.

This is an interesting suggestion. Our long-term goal is to find ways to improve memory, and to better understand the “rules” that might govern changes with age. In this case, were interested in addressing the hypothesis that genes that rise with age must be compensatory, which is a frequently stated theory that is not often tested. Here we showed that knocking down three genes that are upregulated in aged animals improved memory; our results suggest that the wild-type functions of these genes are likely deleterious for learning and memory functions, and further, that their increased expression with age is not a compensatory function. Certainly for future work, it might be interesting to better understand how and why these specific genes have a deleterious function that increases with age, and whether that function is different in younger animals where they are not highly expressed.

The authors then characterize learning and memory in wild-type, daf-2, and daf-2/daf-16 worms with age and find that daf-2 worms have an extended ability to learn for approximately 10 days longer than wild types. This was daf-16 dependent. Memory was extended in daf-2 as well, and strikingly, daf-2;daf-16 had no short-term memory even at day 1. Transcriptomic analysis of FACS-sorted neurons was performed on the three groups at day 8. The authors focus their analysis on daf-2 vs. daf-2;daf-16 and present evidence that daf-2 neurons express a stress-resistance gene program. One question that remains unanswered is how well the N2 and daf-2;daf-16 correlate overall, and are there differences? This may be informative as wild type and daf-2;daf-16 mutants are not phenotypically identical when it comes to memory, and there may be differences that can be detected despite the overlap in the PCA. This analysis could reveal the daf-16 targets involved in memory.

Re. daf-2;daf-16 vs N2: This is a good suggestion. Our analysis in Fig. S5 showed that the daf-2 vs N2 comparison shows similar results with the daf-2 vs daf-16;daf-2 comparison, but some additional genes are differentially expressed. Interestingly, the daf-2 vs N2 comparison shows that the bZip transcription factors are upregulated in daf-2 compared with N2 worms (Fig. S6f). This may indicate that additional transcription factors are controlled by the daf-2 mutation in the nervous system in addition to the DAF-16/FOXO transcription factor.

We also identified the differentially expressed genes in the Day 8 neuronal daf-16;daf-2 to N2 comparison, as the reviewer is asking about. The samples from different genotypes do separate from one another in the PCA plot, indicating there are differences between daf-16,daf-2 and N2 neurons. However, the difference is smaller and there are fewer genes differentially expressed between daf-16;daf-2 and N2: only 38 genes are significantly higher in daf-16;daf-2, and only 53 genes are significantly higher in N2 (log2FC > 0.5, p-adj<0.05). The genes higher in N2 are enriched in endopeptidase inhibitors, and the genes higher in daf-16;daf-2 are not enriched in any gene ontology terms. These results indicate that there are some differences between daf-16;daf-2 and N2 neurons, which correlates with the behavioral differences we see, but the difference is small compared to daf-2 neurons. We have added these data to the paper (Fig. S4e,f); thank you for the suggestion.

The authors tested eight candidate genes that were more highly expressed in daf-2 neurons vs. daf-2;daf-16 and showed that reduction of 2 and 5 of these genes impaired learning and memory, respectively, in daf-2 worms. This finding implicates specific neuronal transcriptional targets of IIS in maintaining cognitive ability in daf-2 with age, which, importantly, are distinct from those in young wild type worms.

Reviewer #2 (Public Review):

Weng et al. perform a comprehensive study of gene expression changes in young and old animals, in wild-type and daf-2 insulin receptor mutants, in the whole animal, and specifically in the nervous system. Using this data, they identify gene families that are correlated with neuronal ageing, as well as a distinct set of genes that are upregulated in neurons of aged daf-2 mutants. This is particularly interesting as daf-2 mutants show both extended lifespans and healthier neurons in aged animals, reflected by better learning/memory in older animals compared with wild-type controls. Indeed, the knockdown of several of these upregulated genes resulted in poorer learning and memory. In addition, the authors showed that several genes upregulated during ageing in wild-type neurons also contribute to learning and memory; specifically knockdown of these genes in young animals resulted in improved memory. This indicates that (at least in this small number of cases), genes that show increased transcript levels with age in the nervous system somehow suppress memory, potentially by having damaging effects on neuronal health.

Finally, from a resource perspective, the neuronal transcriptome provided here will be very useful for C. elegans researchers as it adds to other existing datasets by providing the transcriptome of older animals (animals at day 8 of adulthood) and demonstrating the benefits of performing tissue-specific RNAseq instead of whole-animal sequencing.

Thank you!

The work presented here is of high quality and the authors present convincing evidence supporting their conclusions.

Thanks!

I only have a few comments/suggestions:

(1) Do the genes identified to decrease learning/memory capacity in daf-2 animals (Figure 4d/e) also impact neuronal health? daf-2 mutant worms show delayed onset of age-related changes to neuron structure (Tank et al., 2011, J Neurosci). Does knockdown of the genes shown to affect learning also affect neuron structure during ageing, potentially one mechanism through which they modulate learning/memory?

Thank you for this suggestion, which would be good for a future direction, particularly for genes that might have some relationship to previously-identified cellular structural process. The genes we tested here include dod-24, alh-2, mtl-1, F08H9,4, C44B7.5, hsp-12.3, hsp-12.6, and cpi-1, which are related to stress response, proteolysis inhibitor, metabolic, and innate immunity GO categories, thus associated with stress resistance, proteolysis, lipid metabolism processes; none are obvious choices for morphological effects.

However, it is worth noting that learning and memory decline much faster (Days 4-8) than morphological differences are observed (generally after Day 12-15). Moreover, those morphological differences have been studied primarily in mechanosensory neurons (touch neurons) rather than the chemosensory neurons that are involved in learning and memory, so additional genes may be required for those differences that we were not focusing on in thisi study.

(2) The learning and memory assay data presented in this study uses the butanone olfactory learning paradigm, which is well established by the same group. Have the authors tried other learning assays when testing for learning/memory changes after the knockdown of candidate genes? Depending on the expression pattern of these genes, they may have more or less of an effect on olfactory learning versus for example gustatory or mechanosensory-based learning.

The reason that we use the butanone olfactory learning paradigm is because it is more similar to learning of information (neutral odorant association with positive cue (food)) – the kind of memory we would like to preserve in humans - rather than a stress-induced memory, such as starvation or pathogenesis-associated aversive learning paradigms, which are more like PTSD. (There is likely to be quite a bit of overlap in mechanism, however, including the role of genes such as magi-1 and casy-1, so it would not be surprising if many of these genes also were required for other learning paradigms.)

(3) I have a comment on the 'compensatory vs dysregulatory' model as stated by the authors on page 7. I understand that this model presents the two main options, but perhaps this is slightly too simplistic: the gene expression that rises during ageing may be detrimental for memory (= dysregulatory), but at the same time may also be beneficial for other physiological roles in other tissues (=compensatory).

This is a good point, and we made the clarification that in the text: “There may be other scenarios in which a gene with multiple functions may be detrimental for some behaviors but beneficial for other physiological roles.”

Reviewer #3 (Public Review):

Summary:

In this manuscript, Weng et al. detect a neuron-specific transcriptome that regulates aging. The authors first profile neuron-specific responses during aging at a time point where a loss in memory function is present. They discover signatures unique to neurons which validate their pipeline and reveal the loss of neuron identity with age. For example, old neurons reduce the expression of genes related to synaptic function and neuropeptide signaling and increase the expression of chromatin regulators, insulin peptides, and glycoproteins. The authors discover the detrimental effect of selected upregulated genes (utx-1, ins-19, and nmgp-1) by knocking them down in the whole body and detecting improvement of short memory functions. They then use their pipeline to test neuronal profiles of long-lived insulin/IGF mutants. They discover that genes related to stress response pathways are upregulated upon longevity (e.g. dod-24, F08H9.4) and that they are required for improved neuron function in long-lived individuals.

Strengths:

Overall, the manuscript is well-written, and the experiments are well-described. The authors take great care to explain their reasoning for performing experiments in a specific way and guide the reader through the interpretation of the results, which makes this manuscript an enjoyable and interesting read. Using neuron-specific transcriptomic analysis in aged animals the authors discover novel regulators of learning and memory, which underlines the importance of cell-specific deep sequencing. The time points of the transcriptomic profiling are elegantly chosen, as they coincide with the loss of memory and can be used to specifically reveal gene expression profiles related to neuron function. The authors showcase on the dod-24 example how powerful this approach is. In long-lived insulin/IGF-1 receptor mutants body-wide dod-24 expression differs from neuron-specific profiles. Importantly, the depletion of dod-24 has an opposing effect on lifespan and learning memory. The dataset will provide a useful resource for the C. elegans and aging community.

Thank you, we do hope people will find the data useful.

Weaknesses:

While this study nicely describes the neuron-specific profiles, the authors do not test the relevance in a tissue-specific way. It remains unclear if modifying the responses only in neurons has implications for either memory or potentially for lifespan. The authors point to this in the text and refer to tissue-specific datasets. However, it is possible that the tissue-specific profile changes with age. The authors should consider mining publicly available cell-specific aging datasets and performing neuron-specific RNAi to test the functional relevance of the neuron-specific response. This would strengthen the importance of cell-specific profiling.

Thank you for your suggestions. As we have mentioned in the text, our candidate genes are either (1) only expressed in the neurons (alh-2 and F08H9.4), or they are only more highly expressed in daf-2 compared to wild type only in the nervous system (C44B7.5 or dod-24). Thus, the effect we see from knocking down these genes in daf-2 are likely neuron-specific. Additionaly, we performed our assays with neuron-sensitive RNAi strain CQ745: daf-2(e1370) III; vIs69 [pCFJ90(Pmyo-2::mCherry + Punc-119::sid-1)] V. It has been previously shown that neuronal expression of sid-1 decreases non-neuronal RNAi, suggesting that neurons expressing transgenic sid-1(+) served as a sink for dsRNA (Calixto et al., 2010). Thus, this neuron-sensitive RNAi is likely neuron-specific and our results is unlikely from knocking down these genes in non-neuronal tissues. However, we do acknowledge this issue.

To identify the expression pattern of these genes in a more cell-specific way in the adults, we examined the expression of our candidate genes that affected learning and memory, namely dod-24, F08H9.4, C44B7.9, alh-2, and mtl-1, in the Calico database (Roux et al., 2023). From that database, we can see that dod-24 is mainly expressed in the PHC and PVM neurons, and F08H9.4 is largely expressed in various neurons. Both have only slight expression outside the nervous system. C44B7.5 and mtl-1 are more broadly expressed, but C44B7.5 was not found to be differentially expressed in other tissues in daf-2, and mtl-1 only had a slight effect on learning and memory. Perhaps due to their sequencing depth and detection limit, Roux et al. didn’t detect alh-2 expression anywhere in their data.

Thus, the neuron-specific expression and daf-2 differential expression pattern of these genes indicate that the learning and memory improvement in aged daf-2 is unlikely due to neuronal non-autonomous effects.

To better address this concern (that for the genes that we found only expressed in the neurons, the neuron-confined expression may change with age) we examined the expression pattern change of these genes with age. As is shown below, from the Calico database, we can see that the expression in the nervous system persists, and even slightly increases, with age, thus age-related expression pattern change is not a concern to our analysis.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Most of my comments are in the public section. A few additional recommendations for the authors regarding the formatting/presentation:

The presentation of Figure S6e-h in the introduction is somewhat confusing and feels out of order. If presented first, it should be S1. Otherwise, discussion of this figure should go at the end of the results section or in the discussion if appropriate.

Thank you for pointing this out. We have moved the discussion of this figure to the Discussion section.

I do not see Figure S5 described in the text.

Good catch, thank you. We have added the descriptions for Figure S5 in the text.

In general, check the figures, figure legends, and how they are referenced in the text, particularly the supplemental figures and legends.

Minor comments:

There is a typo in the Figure 4 legend: Neuronal IIX should be IIS.

Thanks for pointing this out. We have corrected it in the text.

Reviewer #2 (Recommendations For The Authors):

• There are multiple instances throughout the manuscript where there are statements in brackets that provide justification or explanation for some of the approaches used. There is no reason for 'side note' brackets to be used. I suggest removing them and incorporating these statements into the narrative.

Thank you, we have now incorporated these points into the main text.

• Introduction: page 4 "here we RNA-sequenced FACS-isolated neurons" should be "here we performed RNA sequencing on FACS-isolated neurons...".

Thank you, we have changed the text accordingly.

• Figure 2A: I do not understand the legend for this panel "Tissue Query for wild-type genes expressed at higher levels in aged worms show lower nervous system and neuron prediction score." Please clarify.

We have clarified the Figure 2A legend:

(A) Tissue prediction score for wild-type genes expressed at higher levels in aged worms.

• Page 8: "We previously observed that loss of single genes that play a role in complex behaviors like learning and memory can have a large impact on function 60, unlike the additive roles of longevity-promoting genes 11." - a large impact on what function?

Thank you for noting, we have clarified it in the text accordingly:

“We previously observed that for genes that play a role in complex behaviors like learning and memory, the loss of single genes can have a large impact on these complex behaviors 60, unlike the additive roles of longevity-promoting genes 11.”

• Next line "Therefore, one mechanism by which wild-type worms lose their function with age..." - again, what function?

Thank you for noting this, we have clarified the text to say we refer to the learning and memory functions.

• Page 9: "Thus, daf-2 mutants maintain their higher cognitive quality of life longer than wild-type worms, while daf-16;daf-2 mutants spend their whole lives without memory ability (Figure 3d), in contrast to claims that daf-2 mutants are less healthy than wild-type or daf-16 worms23." - since ref 23 did not perform any learning/memory tests, the definition of 'health' in ref 23 is different to 'cognitive health' as studied here. So the findings in this study are not 'in contrast' to ref 23 but rather add to these findings.

Learning and memory ability is an important function for a healthy individual, thus we would assert that indeed, cognitive health is an important part of the “health” of daf-2 worms. In ref 23, Bansal et al. claim that daf-2 worms are less healthy without assessing their learning and memory ability; their lack of data is an insufficient reason for us to remove our statement, as cognitive health is part of healthspan. Here we find that the “learning span” of daf-2 lasts at least proportionally if not longer than that of wild type. We have also previously shown that daf-2 worms also have longer maximum velocity span with age (Hahm et al., 2015), in direct contrast with Bansal et al.’s claim that daf-2 worms move less well and thus are less healthy – daf-2 worms simply stop sooner when presented with food and switch to feeding, due to their higher odr-10 levels. The Bansal paper continues to be frequently cited as finding that daf-2 mutants are less healthy than wild type, a claim for which we can still find no experimental evidence to support. Therefore, it is important that we make the point that daf-2 worms have extended cognitive health, which is part of health span.

• Page 13: I feel like the sentence "Furthermore, memory maintenance with age might require additional functions that were not previously uncovered in analyses of young animals" is both vague (what functions are referred to?) and a little bit obvious (obvious that age-related changes would not be revealed in analyses of young animals). Perhaps rephrase to make the desired point clearer?

We have clarified the sentence in the text:

“Furthermore, memory maintenance with age might require additional genes that function in promoting stress resistance and neuronal resilience, which were not previously uncovered in analyses of young animals.”

Version 2

Abstract

Abstract

Cognitive decline is a significant health concern in our aging society. Here, we used the model organism C. elegans to investigate the impact of the IIS/FOXO pathway on age-related cognitive decline. The daf-2 Insulin/IGF-1 receptor mutant exhibits a significant extension of learning and memory span with age compared to wild-type worms, an effect that is dependent on the DAF-16 transcription factor. To identify possible mechanisms by which aging daf-2 mutants maintain learning and memory with age while wild-type worms lose neuronal function, we carried out neuron-specific transcriptomic analysis in aged animals. We observed downregulation of neuronal genes and upregulation of transcriptional regulation genes in aging wild-type neurons. By contrast, IIS/FOXO pathway mutants exhibit distinct neuronal transcriptomic alterations in response to cognitive aging, including upregulation of stress response genes and downregulation of specific insulin signaling genes. We tested the roles of significantly transcriptionally-changed genes in regulating cognitive functions, identifying novel regulators of learning and memory. In addition to other mechanistic insights, comparison of the aged vs young daf-2 neuronal transcriptome revealed that a new set of potentially neuroprotective genes is upregulated; instead of simply mimicking a young state, daf-2 may enhance neuronal resilience to accumulation of harm and take a more active approach to combat aging. These findings suggest a potential mechanism for regulating cognitive function with age and offer insights into novel therapeutic targets for age-related cognitive decline.

Introduction

The loss of cognitive function is a rising problem in our aging society. A 2008 study estimated that at least 22.2% (about 5.4 million) individuals over the age of 71 in the United States have at least mild cognitive impairment1–3. Furthermore, global dementia cases are predicted to triple from an estimated 57.4 million cases in 2019 to 152.8 million cases in 20504,5. As most industrialized countries are experiencing a rapid increase in the proportion of the aged population, understanding and potentially preventing the underlying issues of neuronal structural and behavioral decline associated with aging is crucial for societal health.

C. elegans is an excellent model system for studying neuronal aging, given its tractable genetics, short lifespan, and simple nervous system6. Most importantly, C. elegans experiences rapid loss of learning and memory with age7: by Day 4 of adulthood, all long-term associative memory ability is lost, and by Day 8, C. elegans cannot carry out associative learning or short-term associative memory7 – despite the fact that these worms can still move and chemotaxis perfectly well. That is, with age worms first lose long-term memory ability (by Day 4), then short-term memory and learning ability (Day 6-8), then chemotaxis (Day 10-12), then motility (Day 16)7,8. Because learning and memory decline extremely early, we consider worms that are only a week old to already be “cognitively aged,” despite the fact that they can chemotax and move well, and will continue to live for another one to two weeks. Therefore, we can examine neurons from these 7-8 day-old adults to explore the causes of these cognitive declines in animals that are otherwise quite healthy. Many human neuronal aging phenotypes and genes of interest for mammalian neuronal function are conserved in C. elegans9, making discoveries in C. elegans possibly applicable to humans.

The Insulin/IGF-1-like signaling (IIS)/FOXO pathway was first discovered to play a role in longevity in C. elegans. The lifespan of daf-2/Insulin/IGF-1 receptor mutants is twice that of wild-type animals10, and this lifespan extension requires the downstream Forkhead box O (FOXO) transcription factor DAF-1610. DAF-16/FOXO controls the expression of many genes that contribute to longevity, including stress response, proteostasis, autophagy, antimicrobial, and metabolic genes11. As a conserved regulator, the IIS/FOXO pathway also regulates longevity in Drosophila, mice, and humans12–15. In addition to regulating lifespan, the IIS pathway regulates neuronal function via the FOXO transcription factor. In particular, C. elegans IIS/daf-2 mutants display DAF-16-dependent improved learning, short-term memory, and long-term memory7. While both young and old daf-2 adult worms display increased learning and memory relative to wild-type, the duration of this extension is not known, and the mechanisms by which daf-2 mutants maintain neuronal function in older worms is not yet understood. Compared to wild-type worms, daf-2 mutants better maintain maximum velocity8, motility16,17, neuromuscular junctions, the ability to regenerate axons18,19, and neuronal morphology with age20–22. In particular, we previously showed that while daf-2 has lower observed motility on food23, this apparent is due to its high levels of a food receptor, ODR-108, and its downregulation reveals the much higher mobility of daf-2 animals8, even on food, in addition to its much higher and maintained maximum velocity with age. Previously, we found that daf-2 worms also extend learning beyond wild-type’s ability7, but the full duration of this extension with age was not known. That is, exactly how late in life daf-2 mutants can still learn and remember, and whether this is proportional to their lifespan extension, was not previously determined.

We previously performed neuron-specific RNA-sequencing in young (Day 1) adult C. elegans and identified neuron-specific targets24; genes upregulated in daf-2 mutant neurons are distinct from those in the whole animal, and we found that these neuronal genes are necessary for the observed improvements in memory and axon regeneration in daf-2 mutant worms. However, whether daf-2 uses the same or different genes in young and old worms to improve and maintain cognitive function with age is unknown. Recent datasets using whole-animal single-cell RNA-seq have been generated for wild-type and daf-2 worms, and these are sufficient for whole-body aging and pseudobulk analyses25–27, but we have found that those data are not deep enough to use specifically for in-depth analysis of neurons, which can be difficult to gather from whole animals. Other data are from larval stages and cannot be extrapolated to aging adults28. To identify the transcriptional differences in the aging nervous system that might contribute to the loss of neuronal function with age in wild-type worms and the differences responsible for the extended abilities of daf-2 animals, here we performed RNA sequencing on FACS-isolated neurons of aged (Day 8) wild-type and IIS/FOXO mutants. To further investigate the role of the neuronal IIS/FOXO pathway, we identified genes both upregulated by the IIS/FOXO pathway, and genes that are differentially expressed in daf-2 mutants with age. We found that daf-2 differentially-regulated genes in the aged neurons are different from young neurons; in fact, many of these Day 8 daf-2 vs daf-16;daf-2 upregulated genes are stress response and proteolysis genes that may promote neuronal function and health. We then used functional assays to assess the contributions of daf-2-regulated genes to learning and memory. Our results suggest that daf-2’s neuronal targets in older worms are required to maintain neuronal functions with age, suggesting that additional and alternative mechanisms are at work in these aged mutants from their young counterparts.

Results

Wild-type neurons lose their neuronal function and identity with age

Previously, we found that cognitive abilities in C. elegans, including learning, short-term memory, and long-term memory, all decline with age7. Moreover, neuronal morphology and regeneration ability are also impaired with age18,20–22. However, how these phenotypes are regulated at the molecular level in aging neurons remains to be systematically characterized. Therefore, we were interested in first identifying gene expression changes with age in wild-type neurons to characterize the normal physiological aging process. Before choosing timepoints to assess neuronal transcriptome changes, we carried out associative learning and short-term associative memory assays7 as we have previously described7,29–31. Briefly, well-fed worms are starved for one hour, then re-fed while exposed to the neutral odorant butanone for one hour; a choice assay between butanone and control immediately after training tests associative learning, while a choice assay after one hour of recovery on food-only plates tests short-term associative memory7. Adult Day 1 worms are fully developed, young, and healthy, while wild-type Day 7-8 worms, although still in their mid-life, have completely lost their learning and short-term memory abilities already by Day 77 (Fig. 1a), thus we consider them “aged” for the purposes of understanding loss of cognitive ability. Therefore, we reasoned that a comparison of adult wild-type Day 1 neurons with wild-type neurons that are at least aged Day 7-8 should reveal changes with age that result in loss of cognitive function.

Identifying neuronal aging targets in WT worms using neuron-specific RNA-sequencing.

(a) Wild-type learning and 1hr memory results on Day 1 and Day 7. Learning and memory results are represented as learning index (LI). Details of the LI calculation are explained in the methods. Learning, N = 10, memory, N = 5. ****: p < 0.01. Student’s t-test. (b) PCA plot for Day 1 (orange) and Day 8 (blue) neuronal bulk RNA-seq samples. (c) Volcano plot comparing age-associated differentially-expressed genes in WT neurons. Genes downregulated with age (orange) and upregulated with age (blue) were obtained by neuron-specific RNA sequencing of adult wild-type animals with neuron-specific GFP expression. Adjusted p-value < 0.001, log2(Fold-change) > 2. N = 6 biological replicates per age. 1146 genes were significantly downregulated with age (higher in young neurons) and 2016 genes were upregulated with age (higher in old neurons) (d) Tissue prediction scores for genes higher in young neurons. (e) GO terms of genes that decline with age in wild-type neurons. Synaptic and signaling GO terms are enriched in neuronal genes. p-value calculated using hypergeometric distribution probability. (f) Comparison of whole-body higher-in-young genes and neuronal higher-in-young genes. GO Terms and representative genes were performed using g:Profiler software. P-value of overlapping regions were calculated using a hypergeometric calculator. (g) Normalized reads of ins-6, unc-4, mec-7, folt-2, fbf-1, and madd-4, in Day 1 and Day 8 neurons in our dataset. P-adjusted values were calculated from DESeq2 software. Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values.

To identify genes that regulate age-related morphological and functional decline in wild-type neurons, we performed neuron-specific transcriptomic analysis using our previous FACS neuronal isolation method24 on six biological replicates each of Day 1 and Day 8 adult wild-type worms, where 100,000 GFP+ cells were collected for each sample (Fig. S1a-c). Because we previously found that whole-worm analyses mask changes found specifically in neurons24, to complement our aging neuron studies, we also carried out RNA-sequencing analyses of aging whole worms (Fig. S2a-e), which we found is dominated by changes in the extracellular matrix (Fig. S2b), stress response/pathogen genes (Fig. S2c) and the alimentary system (intestine) (Fig. S2d), overshadowing neuronal changes.

Principal components analysis of the FACS-isolated neuron RNA-seq samples indicated that they are well separated by age (Fig. 1b), and downsampling analysis32 (Fig. S1e) suggested that we have sequenced to saturation, with an average of 41,636,463 uniquely-counted reads and detected the expression of 19725 coding and non-coding genes (log10(TPM)>0.5) (Fig. S1d).

Enrichment analysis of genes that are differentially expressed with age (Fig. 1c-e) suggested that neuronal sorting and sequencing were successful because the sequenced genes are enriched for neuronal genes such as mec-7, mec-12, and twk-49, as expected, and less enriched for all other major tissues. Tissue enrichment analysis of differentially-expressed genes suggested that aging neurons lose genes most expressed in the nervous system and neurons (Fig. 1d), as one might expect. Gene ontology (GO) analysis suggested that genes declining with age in neurons encode proteins important in neuronal function (Fig. 1e, f), including synaptic proteins (e.g., srh-59, rab-3, sng-1, sup-1), potassium channels (e.g., egl-23, twk-7, twk-49, ncs-5), and transmembrane transporters (e.g., folt-2, ccb-2, unc-79, exp-1). The decrease in expression of these genes during aging may indicate that neurons are losing their identity and their ability to perform neuronal functions, such as signal transduction and axonal transport, and correlates with the behavioral and morphological declines observed in aging wild-type worms.

Comparing whole-worm sequencing and neuron-specific sequencing (Fig. 1f), we found that genes involved in metabolic processes decline with age only in the body, and genes encoding structural proteins, lipid localization, and muscle system processes decline with age in both the body and in neurons, while neurons specifically lose genes that are associated with neuronal function, including synaptic proteins, neuropeptide signaling, and other neuron functions, correlating with neuronal loss of function with age. Together, these results indicate that neurons harbor many unique age-related changes that could be overshadowed in the whole-worm transcriptome but are revealed by neuron-specific sequencing.

Many genes that are more highly expressed in young neurons are known to be specific to a subset of neurons. ins-6, an insulin-like peptide specific to the ASI, ASJ, and AWA neurons28 that regulates longevity33 and aversive learning34, is significantly downregulated with age (Fig. 1g). srd-23, a serpentine receptor located at the AWB neuron cilia35, also decreases expression with age (Fig. S1f). Furthermore, various genes specific to sensory neurons (txt-12, flp-33), touch neurons (mec-7), and motor neurons (unc-4) decline in expression with age (Fig. 1g, Fig. S1f).

Previous studies showed that loss of genes including ins-634, mec-736, unc-4, folt-219, madd-437, and fbf-131 lead to behavioral dysfunction in motility and chemosensory abilities; therefore, the decreased expression of these neuron type-specific genes with age may impact the function of individual neurons and disrupt neural circuit communication, ultimately contributing to the declines in behavior observed during aging (Fig. 1g).

As neurons age, genes that increase in expression, while assigned to the nervous system (Fig. 2a) are not specific for neuron function; instead, aged wild-type neurons express higher levels of many predicted F-box genes with predicted proteasome E3 activity (e.g., F-box proteins fbxa-158, fbxb-51, pes-2.1, and SKp1-related proteins skr-12, skr-6). Some transcription regulation (e.g., ced-13, tbx-43, nhr-221, end-1), and chromatin structure and function (e.g., his-54, dot-1.2, jmjd-3.2, hil-7, utx-1) genes also increase with age (Fig. 2b), even though neurons appear to lose their neuron-specific transcriptional identity with age.

Genes that increase with age cause behavioral defects.

(a) Tissue prediction score for wild-type genes expressed at higher levels in aged worms. (b) GO terms of genes expressed higher in aged neurons highlight transcription regulation and proteolysis. GO term analysis was done using Wormcat 2.0. (c) Normalized reads of utx-1 on Day 1 and Day 8. (d) Short-term associative memory (STAM) assay shows that adult-only utx-1 knockdown improves 1hr and 2hr memory of wild-type worms on Day 2. (e) Normalized reads of ins-19 on Day 1 and Day 8. (f) ins-19 mutation improves learning and memory in STAM on Day 3 of adulthood. (g) Normalized reads of nmgp-1 on Day 1 and Day 8. (h) nmgp-1 RNAi knockdown improves memory in STAM on Day 2. P-adj value of normalized count change generated from DEseq2 analysis. (c,e,g) Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values. Normalized reads and adjusted p value calculated using the DESeq2 software. Each dot represents one sequencing replicate. (d,f,h) N = 5 plates in each behavioral experiment. Representative result of 2 biological repeats is shown. *: p<0.05. **: p<0.01. ***: p<0.001. ****: p<0.0001. Two-way ANOVA with Tukey’s post-hoc analysis.

One ongoing discussion about changes during aging is how to interpret an increase in expression with age. There are two main models for genes that increase their expression with age and have a resulting impact on function: that they rise with age to compensate for lost function (“compensatory”) and therefore promote function, or that their expression is deleterious to function and only rises with age through dysregulation. If a gene is compensatory, then its knockdown would abrogate learning and memory, even in young animals. If a gene’s function is harmful to neurons, reducing its expression might be beneficial to the worm, even in young animals. (Of course, there may be other scenarios in which a gene with multiple functions may be detrimental for some behaviors but beneficial for other physiological roles.) To test this hypothesis, we reduced expression of a small set of highly-upregulated candidate genes in categories that might function in a compensatory manner.

These include utx-1, a histone demethylase known to play a role in development38 and lifespan in worms39–41, and whose homolog has been implicated in cognition in mammals42,43; ins-19, an insulin-like peptide; and nmgp-1, a neuronal glycoprotein involved in chemosensation44. In each case, we see that gene expression is significantly higher in old than in young neurons (Fig. 2c, e, g). If a gene increases expression to benefit neurons, we would expect to see no difference in memory in young animals where there is no defect; by contrast, if the increase of a gene is deleterious, we would expect to see an improvement in behavior when knocked down, even in young animals. We performed adult-only neuron-specific RNAi knockdown to prevent any possible deleterious effects caused by changes during development, which largely takes place in early larval stages; testing in young adult animals is logical because there is no memory in aged wild-type, so any deleterious effect of knocking down a potentially compensatory gene in an aged would not result in a change. For all behavioral assays, we first prioritized significantly-changed genes with high fold-change, and then those with mammalian homologs.

We found that 48 hrs (L4-Day 2) of adult-only knockdown of utx-1 increases 1hr and 2hr memory (Fig. 2d), the loss-of-function mutation of ins-19 increases both learning and memory (Fig. 2f) and the adult-only knock-down of nmgp-1 extends memory at 2 hours (Fig. 2h). That is, in each of these cases, reduction of these genes did not impair memory, as loss of a compensatory function would appear; rather, loss of these age-upregulated genes improved wild-type memory. These results indicate that at least some neuronal genes that increase with age can have a negative impact on learning and memory, as demonstrated by the improvement of memory when knocked down, even in young animals. While it is still possible that some upregulated genes may act in a compensatory manner, the simplest model is that at least some are actively deleterious for learning and memory. We previously observed that for genes that play a role in complex behaviors like learning and memory, the loss of single genes can have a large impact on these complex behaviors60, unlike the additive roles of longevity-promoting genes 11. Therefore, one mechanism by which wild-type worms lose their learning and memory functions with age is not just by loss of neuronal gene expression, as one might expect, but also by dysregulation of expression of genes that can negatively impact learning and memory.

daf-2 mutants maintain learning and memory with age

We previously found that daf-2 animals have extended motility (Maximum Velocity) that correlates with and predicts their extension of lifespan8. Additionally, not only do young daf-2 worms have better memory than wild-type worms, but daf-2 mutants also maintain learning and memory better with age7,24. However, the duration of this improvement was unknown. To determine the proportion of life that worms can learn and remember, we tested wild-type, daf-2, and daf-16;daf-2 worms for their learning and associative memory ability every day until these functions were lost. We found that while wild-type worms lose their learning and short-term memory abilities by Day 7-8 (Fig. 1a, Fig. 3a, b), learning and memory span were significantly extended in daf-2 mutants (Fig. 3a, b); thus, a comparison of daf-2 neurons with wild-type neurons at Day 8 should reveal differences relevant to cognitive aging. The extension of learning and memory is dependent on the FOXO transcription factor DAF-16 (Fig. 3a); in fact, while daf-16;daf-2 mutants still have the ability to learn for a few days, these mutants are completely unable to carry out any memory ability, even on Day 1. Thus, learning ability, which is similar in wild-type and daf-2;daf-16 mutants, is mechanistically distinct from short-term memory ability31. daf-2 worms maintained learning ability until Day 19 and short-term (1hr) memory ability until Day 15, more than twice the duration of wild-type worms, while daf-16;daf-2 worms exhibit no short-term memory ability, even on Day 1 of adulthood (Fig. 3b). Our data suggest that the learning span-to-lifespan (Fig. S3c) and memory span-to-lifespan ratios in daf-2 worms were similar to or slightly higher than that of wild-type worms (Fig. 3c, d), indicating that daf-2 mutants maintain cognitive function for at least proportionally as long as wild-type worms do. Thus, daf-2 mutants maintain their higher cognitive quality of life longer than wild-type worms, while daf-16;daf-2 mutants spend their whole lives without memory ability (Fig. 3d), in contrast to claims that daf-2 mutants are less healthy than wild-type or daf-16 worms23. Additionally it should be noted that because our choice assays distinguish motility function from learning and memory function29, the improvements in memory with age shown by daf-2 mutants relative to wild type are distinct from daf-2’s improvements in motility that we previously showed8. Therefore, we are interested in these genes that might contribute to the extended cognitive function that daf-2 worms demonstrate.

Identifying neuronal IIS/FOXO targets in aged worms using neuron-specific RNA-sequencing.

(a) daf-2 mutants show better learning maintenance with age compared to N2 and daf-16;daf-2 worms. N = 10 plates in each condition. (b) daf-2 mutants show better memory maintenance with age compared to N2 worms. daf-16;daf-2 worms do not have 1hr memory on Day 1 of adulthood. N = 10 plates in each condition. (c-d) daf-2 mutants have a slightly larger learning span/lifespan ratio and memory span/lifespan ratio than N2 (wild type). Lifespan shown in Figure S3c. (e) PCA plot of Day 8 N2, daf-2, and daf-16;daf-2 neuronal RNA sequencing results. (f) Volcano plot of neuronal daf-2-regulated, daf-16-dependent up- and downregulated genes on adult Day 8 (Adjusted p-value < 0.05, log2(Fold-change) > 0.5, N = 6 biological replicates per strain). 570 genes were significantly upregulated and 814 genes were downregulated in daf-2 neurons compared with daf-16;daf-2. (g) Volcano plot of whole-worm daf-2 vs daf-16;daf-2 differentially-expressed genes during aging. 3154 genes are higher in daf-2, 1289 genes are higher in daf-16;daf-2 (log2[Fold-change(daf-2 vs daf-16;daf-2)] >1.5, p-adjusted <0.01). (h) Comparison of neuronal and whole-worm Day 8 daf-2 differentially-expressed genes (overlap p = 3.34E-63, hypergeometric test). Neuron-specific and shared daf-2 upregulated genes with the highest fold-changes are labeled.

Aging IIS/FOXO neurons express stress-resistance genes to maintain neuronal function with age

To identify genes that may improve memory and slow cognitive aging in long-lived daf-2 mutants, we compared the transcriptional profiles of Day 8 FACS-isolated neurons from daf-2 animals with Day 8 FACS-isolated wild-type and daf-16;daf-2 neurons; by Day 8, wild-type and daf-16;daf-2 worms have already lost their learning and memory ability, but daf-2 worms still maintain their cognitive functions (Fig. 3a, b). It should be noted that wild-type worms still have normal chemotaxis and motility at Day 87, and there is a separation of several days between the loss of cognitive functions and the loss of motility7; therefore, comparison of the neuronal transcriptomes of daf-2 with wild type and daf-16;daf-2 at this age should specifically highlight genes that are required for learning and memory rather than other functions.

The PCA of the daf-2, daf-16;daf-2, and wild-type neuronal Day 8 transcriptomes (Fig. 3e) indicates that aged daf-16;daf-2 mutant neurons are similar to aged wild-type neurons, correlating well with their similarly worsened cognitive functions at this age; that is, at a transcriptomic level, aged (Day 8) wild-type neurons and aged daf-16;daf-2 neurons are similar, which is echoed by their shared inability to carry out learning and memory by Day 8 of adulthood (Fig. 3ea, b). By contrast, the transcriptomes of aged daf-2 mutant neurons are distinct from both aged wild-type and aged daf-16;daf-2 neuron transcriptomes, just as the cognitive abilities of daf-2 are much greater than wild type or daf-16;daf-2 at this age.

Downsampling analysis shows that our sequencing depth is sufficient to saturate the detectable differential expression (Fig. S3g, h). We obtained an average of 47,233,119 counted reads per sample (Supplemental Table 5) and detected the expression of 16,488 coding and non-coding genes (Fig. S3e).

We identified 570 upregulated and 814 downregulated genes in Day 8 daf-2 neurons compared to Day 8 daf-16;daf-2 neurons (Fig. 3f). A large fraction of the downregulated genes in Day 8 daf-2 vs daf-16;daf-2 neurons are “nematode-specific peptide family” (nspc-) genes of unknown function (Fig. 3f). While the daf-2 vs daf-16;daf-2 changes in whole worms largely replicated the results from our previous studies of young animals11,45 (Fig. 3g, Fig S6g), comparison of the daf-2 vs daf-16;daf-2 differential transcriptional changes in Day 8 whole worms and Day 8 neurons reveal shared (174 genes) and neuron-specific gene expression changes (396 genes; dod-24, srh-2, lin-42, etc.) (Fig. 3h, Fig. S4c). Not surprisingly, previously-identified genes from whole-worm daf-2 vs daf-16;daf-2 and N2 (e.g., sod-3, mtl-1, cpi-1, hsp-12.6, etc.) that play roles in both neurons and other tissues even in Day 1 daf-2 mutants appear in the shared list. (Some neuron-specific Day 8 daf-2-upregulated genes have not been reported to be expressed in neurons previously (e.g., spin-2), further suggesting the value of transcriptomic analyses of isolated neurons in mutant backgrounds at this age.)

Many genes upregulated in Day 8 daf-2 neurons relative to daf-16;daf-2 are related to stress responses, including heat stress (e.g., hsp-12.6, hsp-12.3, F08H9.4/hsp), oxidative stress (e.g., sod-3), and metal stress genes (e.g., mtl-1); and proteolysis (e.g., cpi-1, cpr-2, and tep-1). The upregulation of these genes may perform neuroprotective functions, as their homologs in mammals have been shown to do (Table 1). Specifically, 36 of the top 100 upregulated genes have identified orthologs or identified domains with known function, of which 32 of (89%) have functions in promoting neuronal health. These mammalian homologs protect neurons against protein aggregation and harmful metabolites (e.g., cpi-1, alh-2, ttr-41, gpx-5)46–50, maintain synaptic organization and neuronal homeostasis (e.g., dod-24, ptr-19, plep-1)51–53, facilitate neuronal injury repair (e.g., F08H9.4, sod-3)54,55, and maintain normal neuronal function (e.g., lgc-28, slc-36.3, lin-42)56–59. Together, these genes may help maintain daf-2’s neuronal health and protect neurons from accumulation of environmental harm during aging.

We found that about a third of the daf-2-upregulated genes were shared between the daf-2 vs daf-16;daf-2 analysis and the daf-2 vs N2 analysis (338 genes) (Fig. S5). Of the unshared genes, the daf-2-maintained genes that are specific to the daf-2 vs N2 comparison are bZIP transcription factors, including zip-5, zip-4, atf-2, and proteasome components (Figure S6D).

These results indicate that other transcription factors may participate in regulating daf-2 functions in aged neurons in addition to the daf-16/FOXO transcription factor.

IIS/FOXO transcriptomic changes are necessary for daf-2 mutant’s improved neuronal functions

We were interested not only in the genes that remained upregulated with age, but also in genes that might have increased with age in the high-performing daf-2 mutants. That is, are there genes that increase in expression in daf-2 mutants that are necessary or beneficial for their continued high performance with age? Some of the Day 8 daf-2 vs wild-type or daf-16;daf-2 upregulated genes are also Class 1 DAF-16-dependent genes11 (sod-3, hsp-12.3, fat-5, and mtl-1, hil-1, and dao-2). However, many more genes were differentially expressed in Day 8 daf-2 vs daf-16;daf-2 neurons from our Day 1 data24 (Fig. 4a, Fig. S4d). Of the “new” genes – that is, genes upregulated specifically in neurons of Day 8 vs Day 1 of daf-2 vs daf-16;daf-2 – many have mammalian homologs that have been shown to play neuroprotective roles, by protecting against aggregation proteins and harmful metabolites, maintaining synaptic organization, neuronal homoeostasis, or neuronal activity, or facilitating neuronal injury repair (see Table 1 for specific references).

Neuronal IIS/FOXO aging targets regulate memory decline with age in daf-2 worms.

(a) Comparison of neuronal Day 1 and Day 8 daf-2 vs daf-16;daf-2 upregulated genes. All shared genes and top Day 8-specific daf-2 upregulated genes are labeled. (b) daf-2-regulated fold-change profile of candidate genes. All candidates are upregulated in daf-2 mutants. (c) Description of candidate genes. log2(Fold-change) and p-adjusted values from the daf-2 vs daf-16;daf-2 comparison unless stated otherwise. (d) Candidate gene knockdown effects on Day 6 adult daf-2 learning (0hr after conditioning). Two candidate genes, dod-24 and F08H9.4, show a significant decrease in learning ability. N = 5 plates in each condition, merged results of 3 biological repeats shown. (e) Candidate gene knockdown effects on Day 6 adult daf-2 short-term memory (1hr after conditioning). C44B7.5, dod-24, F08H9.4, mtl-1, and alh-2 showed significant decreases in memory. N = 5 plates in each condition, the representative image of 3 biological repeats shown. (d-e) *: p < 0.05. **: p < 0.01. ***: p<0.001. ****: p <0.0001. One-way ANOVA with Dunnet’s post-hoc analysis. Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values.

If the upregulated genes in aged daf-2 neurons are responsible for the extended memory span of daf-2 mutants, knocking down those genes should block older daf-2 mutants’ memory functions. Therefore, we tested the effect of RNAi knockdown of the top fold-change candidate genes on daf-2’s memory in aged adults. We chose Day 6 for testing because by then, like on Day 8, wild-type worms have already lost their learning and most memory abilities, but daf-2 worms retain normal cognitive functions, and this time point avoids the increased naïve chemotaxis that we observe in older daf-2 animals. As shown in Fig. 4b-c, we selected these significantly differentially-expressed candidate genes based on their ranking in fold-change.

Previously, we have found that the top significantly differentially-expressed genes (by fold-change) are most likely to have strong effects on function, while less differentially-changed genes have less of an effect11,24,60, therefore we prioritized genes that are significantly different and the most highly expressed in daf-2 mutants compared to daf-16;daf-2 mutants for subsequent testing (Fig. 4c). daf-2 worms, including neurons, are susceptible to RNA interference24,61. Of the eight candidate genes we tested, the reduction of three of them (F08H9.4, mtl-1, and dod-24, originally classified as a Class II gene with proposed immune activity) significantly decreased daf-2’s learning ability on Day 6 (Fig. 4d). Those genes plus reduction of two additional genes (C44B7.5 and alh-2) affected 1hr memory (Fig. 4e) in Day 6 daf-2 mutants. That is, knockdown of the heat shock-related gene F08H9.4, the innate immunity gene dod-24, aldehyde dehydrogenase alh-2 and previously uncharacterized gene C44B7.5 are required to some degree for daf-2’s extended memory ability. The reduction of the metal stress gene mtl-1, which is expressed in neurons as well as the rest of the body, had a slight effect on learning and memory.

One caveat of these experiments is that, while we found these genes through the isolation of neurons from aged worms and subsequent RNA-sequencing, the knockdown of the genes and its effects are not necessarily neuron-autonomous; however, alh-2 and F08H9.4 were reported to only be expressed in neurons and the cephalic sheath cell62, and C44B7.5 and dod-24, while expressed more broadly, were not upregulated in daf-2 in the whole-worm analysis (Fig. 3f), therefore their effects are most likely neuron-autonomous. In fact, dod-24 is one of the original Class 2 daf-2-downregulated genes from whole-animal analyses, suggesting that dod-24’s increase in expression is specifically in neurons, therefore the effect of its knockdown is most likely to be neuron-autonomous.

Together, these data suggest that the specific genes that are differentially regulated in Day 8 daf-2 mutants may aid in slowing neuronal function decline and behavioral changes associated with aging. Furthermore, memory maintenance with age might require additional genes that function in promoting stress resistance and neuronal resilience, which were not previously uncovered in analyses of young animals.

Discussion

Although it has been shown previously that daf-2 worms maintain various functions with age, how long they can maintain learning and memory with age, and the genes that might be responsible for these extended neuronal functions, have not been previously explored. Here we have found that daf-2 worms maintain learning and memory abilities proportional with (or even slightly beyond) their degree of lifespan extension, underscoring daf-2’s improved healthspan8.

To understand how memory is lost with age and retained in insulin/IGF-1-like signaling mutants, we have characterized the neuronal transcriptomes of aged wild-type worms and IIS (daf-2) and IIS/FOXO (daf-16;daf-2) mutants (Fig. 5). We found that wild-type neuronal aging is characterized by a downregulation of neuronal function genes and an upregulation of proteolysis genes and transcriptional and epigenetic regulators, which together may help explain the loss of neuronal identity and function with age. We also identified the transcriptomic profile accompanying daf-2’s extended learning and memory span. Specifically, daf-2 neurons maintain higher expression of stress response genes and predicted neuronal homeostasis functions (Table 1), which may help make them more resistant to environmental adversities and age-related decline. We also identified genes responsible for wild-type worms’ worsened learning and memory with age.

Aged daf-2 neurons upregulate neuroprotective genes to maintain improved cognitive behaviors.

During normal neuronal aging, neuron-specific genes decrease in expression, while proteolysis and epigenetic regulators are upregulated, resulting in neuron dysfunction and cognitive function loss. In aged daf-2 mutants, upregulation of neuroprotective genes including dod-24, F08H9.4, C44B7.5, alh-2, and mtl-1 contribute to daf-2’s improved cognitive function. The diagram was generated using Biorender.

By employing a FACS-based neuron-sorting technique, we selectively analyzed adult neuron-function-related genes and investigated their aging process, which is not easily discernible through whole-worm sequencing25–27 (Fig. S6). Sequencing many biological repeats of aging neurons to high depth with ribosomal RNA depletion allowed us to detect a larger number of genes compared to other neuron-related bulk and single-cell sequencing profiles26, providing a deep transcriptomic dataset of aged wild-type, IIS mutant, and IIS/FOXO mutant neurons. Our analysis allowed us to identify differentially-expressed genes that are known to be expressed in at a small number of neurons, even for low-abundance genes. Notably, our sequencing results uncovered genes previously not known to be expressed in neurons that remained undetected in other datasets. Moreover, we revealed the involvement of known neuronal genes in the aging process, such as ins-6 and srd-23. We hope that this dataset will become a valuable resource for detecting new candidates in neuronal aging.

For example, dod-24, which we observed to be upregulated in daf-2 neurons and required for daf-2’s extended memory, was downregulated in the daf-2 whole-worm transcriptome (Fig. S4c). dod-24 has been traditionally classified as a Class II gene that is downregulated in daf-2 worms and upregulated by daf-16 RNAi treatment11,45. Functionally, it has been shown to be an innate immunity gene upregulated during pathogen infection63–65, and its whole-body reduction has been shown to extend the lifespan of wild-type animals11. However, here we find that dod-24 is beneficial in the nervous system and required for daf-2’s extended learning and memory in aged worms. This intriguing contrast between the whole-worm transcriptome and the neuron-specific transcriptome suggests that some genes may have distinct regulatory roles in the nervous system, necessitating a more precise approach beyond whole-worm transcriptomics.

Using this neuron-specific sequencing profile of aged cells, we identified key pathways that change during neuron aging. Our sequencing of aged neurons uncovered active transcriptomic alterations during aging, resulting in not just transcriptional silencing but also upregulation of various pathways. We found that the Day 8 daf-2 vs daf-16;daf-2 neuronal differentially-expressed genes that we have newly discovered here differ from the neuronal Day 1 daf-2 vs daf-16;daf-2 dataset we previously obtained24. These Day 8 differentially-expressed genes are not canonical neuronal genes, such as receptors and ion channels; instead, there are more metabolic and proteolytic genes whose protein orthologs have been shown to be neuroprotective. All top 50 genes and 90% of the top 100 genes with identified mammalian orthologs have been shown to be essential to neuronal functions in mammals (Table 1). These results indicate that instead of simply mimicking a young state, daf-2 may enhance neuron’s resilience to accumulation of harm and take a more active approach to combat aging. These changes suggest that daf-2’s extended memory maintenance may require different mechanisms than function in young animals; daf-2 may maintain neuronal function not just by retaining a youthful transcriptome, but also by increasing the expression of genes that promote resilience, such as stress-response genes and proteolysis inhibitors.

In addition to examining aging in wild-type and IIS/FOXO mutants independently, our results further linked the normal aging process to altered gene regulation in the IIS pathway. utx-1, nmgp-1, and ins-19 increase in expression in aged neurons, and we found that their reduction improved memory, indicating that at least some of the genes whose expression rises with age can have a negative impact on normal cognitive functions, rather than acting in a compensatory manner. utx-1 is an H3K27me3 histone demethylase we found to be higher in wild-type aged neurons, but it is also involved in the IIS pathway. The downregulation of utx-1 has been shown to regulate development38 and promote longevity39–41, and its mammalian homolog has been implicated in regulating cognitive abilities42,43. The longevity response of utx-1 depends on daf-1639–41. The loss of utx-1 decreases methylation on the daf-2 gene, thus increasing DAF-16’s nuclear localization, mimicking a daf-2 mutation39. This example of the crosstalk between normal aging and IIS/FOXO mutants offers valuable insights into modifying the aging process for enhanced longevity and cognitive health.

We found that the insulin-like peptide ins-19 was upregulated in aged neurons and was downregulated in aged daf-2 neurons, and its downregulation in wild-type worms extended memory span. Insulin-like peptides play crucial roles as receptor ligands (in both agonist and antagonist roles) for DAF-2, and we have found them to be downregulated in daf-2 mutants compared with daf-16;daf-2 mutants, possibly creating a feedback loop that dampens the insulin signaling pathway, as was previously shown for ins-7 and ins-1811,66. These peptides exhibit diverse functions in development, dauer formation, and longevity11,66–70. Notably, certain insulin-like peptides have been linked to neuronal activities, such as the regulation of aversive learning by the two antagonistic peptides ins-6 and ins-734, and reduced long-term learning and memory by ins-22 RNAi60. In our study, the expression changes of ins-19 during wild-type aging and in daf-2 mutants provide an example of how longevity mutants can reverse wild-type transcriptional changes during aging, ultimately reducing behavioral and functional decline.

Summary of mechanistic insights

Our analysis of transcriptomes from isolated aged Day 8 neurons of wild-type, daf-2, and daf-16;daf-16 mutants reveal several major mechanistic insights. Specifically, we found that wild-type neurons lose their neuronal identity through a combination of the loss of neuron-specific function genes with age, and the concomitant dysregulated increase in non-neuronal genes with age. Furthermore, at least a fraction of the top-upregulated genes with age can play deleterious roles; that is, they rise with age, and their knockdown improves function, even in young animals. This argues against the idea that all of these genes are playing a compensatory role with age.

We also found that the knockdown of individual top-ranked genes that function in learning and memory can have a large impact - like removing a cog of a machine. This is in contrast to our earlier findings regarding gene reduction in lifespan, where most cellular longevity processes regulated by DAF-16 activity appear to be additive, and therefore loss of individual major genes downstream of DAF-2 and DAF-16 have at most a 5-10% impact11. Several of these genes we found to be required for daf-2’s age-related improvement in learning and memory - namely dod-24, F08H9.4, C44B7.5, and alh-2 – were previously not associated with memory function. Finally, these genes are distinct from the set of upregulated Day 1 daf-2 vs daf-16;daf-2 genes; how they each individually maintain neuronal function better with age will be interesting to dissect.

Conclusions

Beyond our sequencing analysis, we have established links between genomics, function, and behavior. We also identified several new genes required for daf-2’s age-related improvement in learning and memory, shedding light on their neuron-specific roles. These additional findings further suggest that neuronal sequencing datasets can be used to identify functional candidates and pathways during the aging process. By bridging the gap between transcriptomic landscapes, genetic regulation, and functional outcomes, our study provides a greater understanding of the mechanisms underlying neuronal aging, providing insights into development of aging interventions.

Methods

Strains and worm cultivation

N2 (wild type), OH441: otIs45(unc-119::GFP), CQ295: otIs45(unc-119::GFP);daf-2(e1370), CQ296: otIs45(unc-119::GFP) ;daf-16(mu86);daf-2(e1370), LC108: uIs69 (myo-2p::mCherry + unc-119p::sid-1), CQ705: daf-2(e1370) III, 3X outcrossed, CQ745: daf-2(e1370) III; vIs69 [pCFJ90(Pmyo-2::mCherry + Punc-119::sid-1)] V, QL188: ins-19(tm5155) II, CX3695: kyIs140(str-2::GFP + lin-15(+)), CQ461: (daf-2(e1370);Pmec-4::mCherry) and CQ501: (daf-2 (e1370);zip-5(gk646);Pmec-4::mCherry). Strains were grown on high-growth media (HGM) plates seeded with E. coli OP50 bacteria using standard methods 71.

Tissue-specific isolation

For neuronal isolation, five plates of fully-grown worms from HG plates were synchronized by hypochlorite treatment, eggs spread on seeded plates to hatch, and at least five plates/replicate were grown to L4 on HGM plates until transferred to HGM plates with FUdR to avoid progeny contamination. This gives us ∼6000 healthy Day 8 worms to sort. Neuron isolation and Fluorescent-activated cell sorting were carried out as previously described 24,62.

Briefly, worms were treated with 1000uL lysis buffer (200 mM DTT, 0.25% SDS, 20 mM HEPES pH 8.0, 3% sucrose) for 6.5 mins to break the cuticle. Then worms were washed and resuspended in 500uL 20mg/mL pronase from Streptomyces griseus (Sigma-Aldrich). Worms were incubated at room temperature with mechanical disruption by pipetting until no whole-worm bodies were seen, and then ice-cold osmolarity adjusted L-15 buffer(Gibco) with 2% Fetal Bovine Serum (Gibco) were added to stop the reaction. Prior to sorting, cell suspensions were filtered using a 5um filter and sorted using a FACSVantage SE w/ DiVa (BD Biosciences; 488nm excitation, 530/30nm bandpass filter for GFP detection). Sorting gates were determined by comparing with age-matched, genotype-matched non-fluorescent cell suspension samples. Fluorescent neuron cells were directly sorted into Trizol LS. 100,000 GFP+ cells were collected for each sample.

RNA extraction, library generation and sequencing

We used standard trizol-chloroform-isopropanol method to extract RNA, then performed RNA cleanup using RNeasy MinElute Cleanup Kit (Qiagen). RNA quality was assessed using the Agilent Bioanalyzer RNA Pico chip, and bioanalyzer RIN>6.0 samples were observed before library generation. 2ng of RNA was used for library generation using Ovation SoLo RNA-Seq library preparation kit with AnyDeplete Probe Mix-C. elegans (Tecan Genomics) according to the manufacturer’s instructions 72. Library quality and concentration was assessed using Agilent Bioanalyzer DNA 12000 chip. Samples were multiplexed and sequencing were performed using NovaSeq S1 100nt Flowcell v1.5 (Illumina).

Data processing

FastQC was performed on each sample for quality control analysis. RNA STAR package was used for mapping paired-end reads to the C. elegans genome ce11 (UCSC Feb 2013) using the gene model ws245genes.gtf. Length of the genomic sequence around annotated junctions is chosen as read length -1. 50-70% reads were uniquely mapped. Reads uniquely mapped to the genome were then counted using htseq-count (mode = union). DESeq2 analysis was then used for read normalization and differential expression analysis on counted reads 73. Genes with a log10TPM >0.5 were considered as detected and genes with a log2(fold-change) > 0.5 and p-adjusted <0.05 are considered differentially expressed in further analysis. Gene ontology analysis were performed using gprofiler74 or WormCat 2.075 and category 2 was selected to show. Tissue query was performed on top 500 highest fold-change genes, using worm tissue query website (www.worm.princeton.edu)62, and only major systems were selected in the analysis.

Learning and Memory Experiments

We performed Short-Term Associative Memory (STAM) experiments as previously described 7,29. Briefly, we used five plates of synchronized adult worms/sample to perform each experiment. One plate of worms was used to test naïve chemotaxis assay without conditioning, while the other 3 plates were washed into M9, and washed 3 additional times to get rid of the bacteria. These worms are starved for 1hr to prime them for food uptake. Then these worms are transferred to conditioning plates with NGM plates seeded with OP50 and 10% butanone stripes on the lid for 1hr to perform conditioning. After conditioning, worms are either transferred from the conditioning plate directly to the chemotaxis plates to assess learning, or transferred to holding plate for 1hr or 2hrs to assess for memory. After staying on holding plates for 1hr or 2hrs, worms are then transferred onto chemotaxis plates to assess for short-term memory. Chemotaxis assays were performed by transferring worms onto chemotaxis plates with 1uL 10% butanone and 1uL ethanol spots separated by 8cm on a 10cm NGM plate. Worms who have reached either butanone spot or the ethanol spot are paralyzed by the 1uL 7.5% NaN3 on these spots. For each timepoint, 5 chemotaxis plates are used to minimize the variation of the outcome. We performed this chemotaxis assay to butanone on naïve and appetitive-trained worms at different time points to assess change in preference to butanone. Chemotaxis index is calculated as (# of worms at butanone-# of worms at ethanol)/(total # of worms - # of worms at origin).

Learning index is calculated by subtracting trained chemotaxis index with naïve chemotaxis index.

For learning and memory span assays, we obtained synchronized worms from hypochlorite-treated eggs. Synchronized worms were washed onto 5’-fluorodeoxyuridine (FUdR) at L4 and maintained on FUdR plates by transferring to new plates every 2 days. 1 Day Prior to experiments, worms are washed onto fresh HG plates without FUdR to avoid change in behavior caused by FUdR. To verify that FUdR has no effect on short-term memory, we compared worms with and without FUdR (Fig. S6d), and found no differences. For utx-1 and nmgp-1 RNAi experiments, synchronized L4 wild-type worms were washed onto HGM plates with carbenicillin and IPTG and seeded with HT115 RNAi bacteria containing the RNAi constructs from the Ahringer Library. For daf-2 upregulated candidates’ RNAi experiments, synchronized L4 daf-2 RNAi-sensitive worms were washed onto HGM plates added with carbenicillin, FUdR, and isopropyl-b-D-thiogalactopyranoside (IPTG) and seeded with HT115 bacteria containing RNAi constructs generated from the Ahringer RNAi Library, then were transferred onto fresh RNAi plates every 2 days until Day 6. 1 Day Prior to experiments, worms are transferred onto plates without FUdR.

Quantitative and Statistical Analysis

All experimental analysis was performed using Prism 8 software. Two-way ANOVA with Tukey post-hoc tests were used to compare the learning curve between control and experimental groups. One-way ANOVA followed by Dunnet post-hoc tests for multiple comparisons was performed to compare learning or 2hr memory between various treatment groups and control. Chi-square test was performed to compare the neuron morphology change between young and aged AWC neurons. All GO term analysis were perform using Wormcat 2.0 software with Bonferroni corrected adjusted p-values. Venn diagram overlaps were compared using the hypergeometric test. Differential expression analysis of RNA-seq were performed using DESeq2 algorithm and adjusted p-values were generated with Wald test using Benjamini and Hochberg method (BH-adjusted p values). Additional statistical details of experiments, including sample size (with n representing the number of chemotaxis assays performed for behavior, RNA collections for RNA-seq, and the number of worms for microscopy), can be found in the methods and figure legends. Regression analyses were performed using sklearn packages. Correlations were calculated using the SciPy packages.

Data availability

Sequencing reads are deposited at NCBI BioProject under accession number PRJNA999305.

Code availability

Analysis codes generated from this study are available upon request.

Resource availability

Further information and requests for resources and reagents should be directed to and will be fulfilled by Coleen T. Murphy (ctmurphy@princeton.edu).

Supporting information

eLife Assessment

This fundamental study investigates the transcriptional changes in neurons that underlie loss of learning and memory with age in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants. The presented evidence is compelling, utilizing a cutting-edge method to isolate neurons from worms for genomics that is clearly conveyed with a rigorous experimental approach. Overall, this study supports that older daf-2 worms maintain cognitive function via mechanisms that are unique from younger wild type worms, which will be of great interest to neuroscientists and researchers studying ageing.

Reviewer #1 (Public Review):

The authors perform RNA-seq on FACS isolated neurons from adult worms at days 1 and 8 of adulthood to profile the gene expression changes that occur with cognitive decline. Supporting data are included indicating that by day 7 of adulthood, learning and memory are reduced, indicating that this timepoint or after represents cognitively aged worms. Neuronal identity genes are reduced in expression within the cognitively aged worms, whereas genes involved in proteostasis, transcription/chromatin, and the stress response are elevated. A number of specific examples are provided, representing markers of specific neuronal subtypes, and correlating expression changes to the erosion of particular functions (e.g. motor neurons, chemosensory neurons, aversive learning neurons, etc).

To investigate whether upregulation of genes in neurons with age is compensatory or deleterious, the authors reduced expression of a set of three significantly upregulated genes and performed behavioral assays in young adults. In each case, reduction of expression improved memory, consistent with a model in which age-associated increases impair neuronal function.

The authors then characterize learning and memory in wild type, daf-2, and daf-2/daf-16 worms with age and find that daf-2 worms have an extended ability to learn for approximately 10 days longer that wild types. This was daf-16 dependent. Memory was extended in daf-2 as well, and strikingly, daf-2;daf-16 had no short term memory even at day 1. Transcriptomic analysis of FACS-sorted neurons was performed on the three groups at day 8. The authors focus their analysis on daf-2 vs. daf-2;daf-16 and present evidence that daf-2 neurons express a stress-resistance gene program. They also find small differences between the N2 and daf-2;daf-16 neurons, which correlate with the observed behavioral differences, though these differences are modest.

The authors tested eight candidate genes that were more highly expressed in daf-2 neurons vs. daf-2;daf-16 and showed that reduction of 2 and 5 of these genes impaired learning and memory, respectively, in daf-2 worms. This finding implicates specific neuronal transcriptional targets of IIS in maintaining cognitive ability in daf-2 with age, which, importantly, are distinct from those in young wild type worms.

Overall, this is a strong study with rigorously performed experiments. The authors achieved their aim of identifying transcriptional changes in neurons that underlie loss of learning and memory in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants.

Reviewer #2 (Public Review):

Weng et al. perform a comprehensive study of gene expression changes in young and old animals, in wild-type and daf-2 insulin receptor mutants, in the whole animal and specifically in the nervous system. Using this data, they identify gene families that are correlated with neuronal ageing, as well as a distinct set of genes that are upregulated in neurons of aged daf-2 mutants. This is particularly interesting as daf-2 mutants show both extended lifespan and healthier neurons in aged animals, reflected by better learning/memory in older animals compared with wild-type controls. Indeed, knockdown of several of these upregulated genes resulted in poorer learning and memory. In addition, the authors showed that several genes upregulated during ageing in wild-type neurons also contribute to learning and memory; specifically, knockdown of these genes in young animals resulted in improved memory. This indicates that (at least in this small number of cases), genes that show increased transcript levels with age in the nervous system somehow suppress memory, potentially by having damaging effects on neuronal health.

Finally, from a resource perspective, the neuronal transcriptome provided here will be very useful for C. elegans researchers as it adds to other existing datasets by providing the transcriptome of older animals (animals at day 8 of adulthood) and demonstrating the benefits of performing tissue-specific RNAseq instead of whole-animal sequencing.

The work presented here is of high quality and the authors present convincing evidence supporting their conclusions. I only have a few comments/suggestions:

(1) Do the genes identified to decrease learning/memory capacity in daf-2 animals (Figure 4d/e) also impact neuronal health? daf-2 mutant worms show delayed onset of age-related changes to neuron structure (Tank et al., 2011, J Neurosci). Does knockdown of the genes shown to affect learning also affect neuron structure during ageing, potentially one mechanism through which they modulate learning/memory?

(2) The learning and memory assay data presented in this study uses the butanone olfactory learning paradigm, which is well established by the same group. Have the authors tried other learning assays when testing for learning/memory changes after knockdown of candidate genes? Depending on the expression pattern of these genes, they may have more or less of an effect on olfactory learning versus for e.g. gustatory or mechanosensory-based learning.

(3) A comment on the 'compensatory vs dysregulatory' model as stated by the authors on page 7 - I understand that this model presents the two main options, but perhaps this is slightly too simplistic: gene expression that rises during ageing may be detrimental for memory (= dysregulatory), but at the same time may also be beneficial other physiological roles in other tissues (=compensatory).

Comments on revised version:

I am satisfied with how the authors have addressed all my comments/suggestions.

Reviewer #3 (Public Review):

Summary

In this manuscript, Weng et al. identify the neuron specific transcriptome that impacts age dependent cognitive decline. The authors design a pipeline to profile neurons from wild type and long-lived insulin receptor/IGF-1 mutants using timepoints when memory functions are declining. They discover signatures unique to neurons which validates their approach. The authors identify that genes related to neuronal identity are lost with age in wild type worms. For example, old neurons reduce the expression of genes linked to synaptic function and neuropeptide signaling and increase the expression of chromatin regulators, insulin peptides and glycoproteins. Depletion of selected genes which are upregulated in old neurons (utx-1, ins-19 and nmgp-1) leads to improved short memory function. This indicates that some genes that increase with age have detrimental effects on learning and memory. The pipeline is then used to test neuronal profiles of long-lived insulin/IGF-1 daf-2 mutants. Genes related to stress response pathways are upregulated in long lived daf-2 mutants (e.g. dod-24, F08H9.4) and those genes are required for improved neuron function.

Strengths

The manuscript is well written, and the experiments are well described. The authors take great care to explain their reasoning for performing experiments in a specific way and guide the reader through the interpretation of the results, which makes this manuscript an enjoyable and interesting read. The authors discover novel regulators of learning and memory using neuron-specific transcriptomic analysis in aged animals, which underlines the importance of cell specific deep sequencing. The timepoints of the transcriptomic profiling are elegantly chosen, as they coincide with the loss of memory and can be used to specifically reveal gene expression profiles related to neuron function. The authors discuss on the dod-24 example how powerful this approach is. In daf-2 mutants whole-body dod-24 expression differs from neuron specific profiles, which underlines the importance of precise cell specific approaches. This dataset will provide a very useful resource for the C. elegans and aging community as it complements existing datasets with additional time points and neuron specific deep profiling.

Weakness

This study nicely describes the neuron specific profiles of aged long-lived daf-2 mutants. Selected neuronal genes that were upregulated in daf-2 mutants (e.g. F08H9.4, mtl-1, dod-24, alh-2, C44B7.5) decreased learning/memory when knocked down. However, the knock down of these genes was not specific to neurons. The authors use a neuron-sensitive RNAi strain to address this concern and acknowledge this caveat in the text. While it is likely that selected candidates act only in neurons it is possible that other tissues participate as well.

Author response:

The following is the authors’ response to the previous reviews.

eLife assessment

This fundamental study investigates the transcriptional changes in neurons that underlie loss of learning and memory with age in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants. The presented evidence is compelling, utilizing a cutting-edge method to isolate neurons from worms for genomics that is clearly conveyed with a rigorous experimental approach. Overall, this study supports that older daf-2 worms maintain cognitive function via mechanisms that are unique from younger wild type worms, which will be of great interest to neuroscientists and researchers studying ageing.

Public Reviews:

Reviewer #1 (Public Review):

The authors perform RNA-seq on FACS isolated neurons from adult worms at days 1 and 8 of adulthood to profile the gene expression changes that occur with cognitive decline. Supporting data are included indicating that by day 7 of adulthood, learning and memory are reduced, indicating that this timepoint or after represents cognitively aged worms. Neuronal identity genes are reduced in expression within the cognitively aged worms, whereas genes involved in proteostasis, transcription/chromatin, and the stress response are elevated. A number of specific examples are provided, representing markers of specific neuronal subtypes, and correlating expression changes to the erosion of particular functions (e.g. motor neurons, chemosensory neurons, aversive learning neurons, etc).

To investigate whether upregulation of genes in neurons with age is compensatory or deleterious, the authors reduced expression of a set of three significantly upregulated genes and performed behavioral assays in young adults. In each case, reduction of expression improved memory, consistent with a model in which age-associated increases impair neuronal function.

The authors then characterize learning and memory in wild type, daf-2, and daf-2/daf-16 worms with age and find that daf-2 worms have an extended ability to learn for approximately 10 days longer that wild types. This was daf-16 dependent. Memory was extended in daf-2 as well, and strikingly, daf-2;daf-16 had no short term memory even at day 1. Transcriptomic analysis of FACS-sorted neurons was performed on the three groups at day 8. The authors focus their analysis on daf-2 vs. daf-2;daf-16 and present evidence that daf-2 neurons express a stress-resistance gene program. They also find small differences between the N2 and daf-2;daf-16 neurons, which correlate with the observed behavioral differences, though these differences are modest.

The authors tested eight candidate genes that were more highly expressed in daf-2 neurons vs. daf-2;daf-16 and showed that reduction of 2 and 5 of these genes impaired learning and memory, respectively, in daf-2 worms. This finding implicates specific neuronal transcriptional targets of IIS in maintaining cognitive ability in daf-2 with age, which, importantly, are distinct from those in young wild type worms.

Overall, this is a strong study with rigorously performed experiments. The authors achieved their aim of identifying transcriptional changes in neurons that underlie loss of learning and memory in C. elegans, and how cognition is maintained in insulin/IGF-1-like signaling mutants.

We thank you for the evaluation and response.

Reviewer #2 (Public Review):

Weng et al. perform a comprehensive study of gene expression changes in young and old animals, in wild-type and daf-2 insulin receptor mutants, in the whole animal and specifically in the nervous system. Using this data, they identify gene families that are correlated with neuronal ageing, as well as a distinct set of genes that are upregulated in neurons of aged daf-2 mutants. This is particularly interesting as daf-2 mutants show both extended lifespan and healthier neurons in aged animals, reflected by better learning/memory in older animals compared with wild-type controls. Indeed, knockdown of several of these upregulated genes resulted in poorer learning and memory. In addition, the authors showed that several genes upregulated during ageing in wild-type neurons also contribute to learning and memory; specifically, knockdown of these genes in young animals resulted in improved memory. This indicates that (at least in this small number of cases), genes that show increased transcript levels with age in the nervous system somehow suppress memory, potentially by having damaging effects on neuronal health.

Finally, from a resource perspective, the neuronal transcriptome provided here will be very useful for C. elegans researchers as it adds to other existing datasets by providing the transcriptome of older animals (animals at day 8 of adulthood) and demonstrating the benefits of performing tissue-specific RNAseq instead of whole-animal sequencing.

The work presented here is of high quality and the authors present convincing evidence supporting their conclusions. I only have a few comments/suggestions:

(1) Do the genes identified to decrease learning/memory capacity in daf-2 animals (Figure 4d/e) also impact neuronal health? daf-2 mutant worms show delayed onset of age-related changes to neuron structure (Tank et al., 2011, J Neurosci). Does knockdown of the genes shown to affect learning also affect neuron structure during ageing, potentially one mechanism through which they modulate learning/memory?

(2) The learning and memory assay data presented in this study uses the butanone olfactory learning paradigm, which is well established by the same group. Have the authors tried other learning assays when testing for learning/memory changes after knockdown of candidate genes? Depending on the expression pattern of these genes, they may have more or less of an effect on olfactory learning versus for e.g. gustatory or mechanosensory-based learning.

(3) A comment on the 'compensatory vs dysregulatory' model as stated by the authors on page 7 - I understand that this model presents the two main options, but perhaps this is slightly too simplistic: gene expression that rises during ageing may be detrimental for memory (= dysregulatory), but at the same time may also be beneficial other physiological roles in other tissues (=compensatory).

Thank you for your original suggestions; we addressed them in the previous version of response to the reviewers.

Comments on revised version:

I am satisfied with how the authors have addressed all my comments/suggestions.

Thank you for your response!

Reviewer #3 (Public Review):

Summary

In this manuscript, Weng et al. identify the neuron specific transcriptome that impacts age dependent cognitive decline. The authors design a pipeline to profile neurons from wild type and long-lived insulin receptor/IGF-1 mutants using timepoints when memory functions are declining. They discover signatures unique to neurons which validates their approach. The authors identify that genes related to neuronal identity are lost with age in wild type worms. For example, old neurons reduce the expression of genes linked to synaptic function and neuropeptide signaling and increase the expression of chromatin regulators, insulin peptides and glycoproteins. Depletion of selected genes which are upregulated in old neurons (utx-1, ins-19 and nmgp-1) leads to improved short memory function. This indicates that some genes that increase with age have detrimental effects on learning and memory. The pipeline is then used to test neuronal profiles of long-lived insulin/IGF-1 daf-2 mutants. Genes related to stress response pathways are upregulated in long lived daf-2 mutants (e.g. dod-24, F08H9.4) and those genes are required for improved neuron function.

Strengths

The manuscript is well written, and the experiments are well described. The authors take great care to explain their reasoning for performing experiments in a specific way and guide the reader through the interpretation of the results, which makes this manuscript an enjoyable and interesting read. The authors discover novel regulators of learning and memory using neuron-specific transcriptomic analysis in aged animals, which underlines the importance of cell specific deep sequencing. The timepoints of the transcriptomic profiling are elegantly chosen, as they coincide with the loss of memory and can be used to specifically reveal gene expression profiles related to neuron function. The authors discuss on the dod-24 example how powerful this approach is. In daf-2 mutants whole-body dod-24 expression differs from neuron specific profiles, which underlines the importance of precise cell specific approaches. This dataset will provide a very useful resource for the C. elegans and aging community as it complements existing datasets with additional time points and neuron specific deep profiling.

Weakness

This study nicely describes the neuron specific profiles of aged long-lived daf-2 mutants. Selected neuronal genes that were upregulated in daf-2 mutants (e.g. F08H9.4, mtl-1, dod-24, alh-2, C44B7.5) decreased learning/memory when knocked down. However, the knock down of these genes was not specific to neurons. The authors use a neuron-sensitive RNAi strain to address this concern and acknowledge this caveat in the text. While it is likely that selected candidates act only in neurons it is possible that other tissues participate as well.

Thank you for pointing this caveat out. We have mentioned it in the figure legend.

Version 3

Abstract

Abstract

Cognitive decline is a significant health concern in our aging society. Here, we used the model organism C. elegans to investigate the impact of the IIS/FOXO pathway on age-related cognitive decline. The daf-2 Insulin/IGF-1 receptor mutant exhibits a significant extension of learning and memory span with age compared to wild-type worms, an effect that is dependent on the DAF-16 transcription factor. To identify possible mechanisms by which aging daf-2 mutants maintain learning and memory with age while wild-type worms lose neuronal function, we carried out neuron-specific transcriptomic analysis in aged animals. We observed downregulation of neuronal genes and upregulation of transcriptional regulation genes in aging wild-type neurons. By contrast, IIS/FOXO pathway mutants exhibit distinct neuronal transcriptomic alterations in response to cognitive aging, including upregulation of stress response genes and downregulation of specific insulin signaling genes. We tested the roles of significantly transcriptionally-changed genes in regulating cognitive functions, identifying novel regulators of learning and memory. In addition to other mechanistic insights, comparison of the aged vs young daf-2 neuronal transcriptome revealed that a new set of potentially neuroprotective genes is upregulated; instead of simply mimicking a young state, daf-2 may enhance neuronal resilience to accumulation of harm and take a more active approach to combat aging. These findings suggest a potential mechanism for regulating cognitive function with age and offer insights into novel therapeutic targets for age-related cognitive decline.

Introduction

The loss of cognitive function is a rising problem in our aging society. A 2008 study estimated that at least 22.2% (about 5.4 million) individuals over the age of 71 in the United States have at least mild cognitive impairment1–3. Furthermore, global dementia cases are predicted to triple from an estimated 57.4 million cases in 2019 to 152.8 million cases in 20504,5. As most industrialized countries are experiencing a rapid increase in the proportion of the aged population, understanding and potentially preventing the underlying issues of neuronal structural and behavioral decline associated with aging is crucial for societal health.

C. elegans is an excellent model system for studying neuronal aging, given its tractable genetics, short lifespan, and simple nervous system6. Most importantly, C. elegans experiences rapid loss of learning and memory with age7: by Day 4 of adulthood, all long-term associative memory ability is lost, and by Day 8, C. elegans cannot carry out associative learning or short-term associative memory7 – despite the fact that these worms can still move and chemotaxis perfectly well. That is, with age worms first lose long-term memory ability (by Day 4), then short-term memory and learning ability (Day 6-8), then chemotaxis (Day 10-12), then motility (Day 16)7,8. Because learning and memory decline extremely early, we consider worms that are only a week old to already be “cognitively aged,” despite the fact that they can chemotax and move well, and will continue to live for another one to two weeks. Therefore, we can examine neurons from these 7-8 day-old adults to explore the causes of these cognitive declines in animals that are otherwise quite healthy. Many human neuronal aging phenotypes and genes of interest for mammalian neuronal function are conserved in C. elegans9, making discoveries in C. elegans possibly applicable to humans.

The Insulin/IGF-1-like signaling (IIS)/FOXO pathway was first discovered to play a role in longevity in C. elegans. The lifespan of daf-2/Insulin/IGF-1 receptor mutants is twice that of wild-type animals10, and this lifespan extension requires the downstream Forkhead box O (FOXO) transcription factor DAF-1610. DAF-16/FOXO controls the expression of many genes that contribute to longevity, including stress response, proteostasis, autophagy, antimicrobial, and metabolic genes11. As a conserved regulator, the IIS/FOXO pathway also regulates longevity in Drosophila, mice, and humans12–15. In addition to regulating lifespan, the IIS pathway regulates neuronal function via the FOXO transcription factor. In particular, C. elegans IIS/daf-2 mutants display DAF-16-dependent improved learning, short-term memory, and long-term memory7. While both young and old daf-2 adult worms display increased learning and memory relative to wild-type, the duration of this extension is not known, and the mechanisms by which daf-2 mutants maintain neuronal function in older worms is not yet understood. Compared to wild-type worms, daf-2 mutants better maintain maximum velocity8, motility16,17, neuromuscular junctions, the ability to regenerate axons18,19, and neuronal morphology with age20–22. In particular, we previously showed that while daf-2 has lower observed motility on food23, this apparent is due to its high levels of a food receptor, ODR-108, and its downregulation reveals the much higher mobility of daf-2 animals8, even on food, in addition to its much higher and maintained maximum velocity with age. Previously, we found that daf-2 worms also extend learning beyond wild-type’s ability7, but the full duration of this extension with age was not known. That is, exactly how late in life daf-2 mutants can still learn and remember, and whether this is proportional to their lifespan extension, was not previously determined.

We previously performed neuron-specific RNA-sequencing in young (Day 1) adult C. elegans and identified neuron-specific targets24; genes upregulated in daf-2 mutant neurons are distinct from those in the whole animal, and we found that these neuronal genes are necessary for the observed improvements in memory and axon regeneration in daf-2 mutant worms. However, whether daf-2 uses the same or different genes in young and old worms to improve and maintain cognitive function with age is unknown. Recent datasets using whole-animal single-cell RNA-seq have been generated for wild-type and daf-2 worms, and these are sufficient for whole-body aging and pseudobulk analyses25–27, but we have found that those data are not deep enough to use specifically for in-depth analysis of neurons, which can be difficult to gather from whole animals. Other data are from larval stages and cannot be extrapolated to aging adults28. To identify the transcriptional differences in the aging nervous system that might contribute to the loss of neuronal function with age in wild-type worms and the differences responsible for the extended abilities of daf-2 animals, here we performed RNA sequencing on FACS-isolated neurons of aged (Day 8) wild-type and IIS/FOXO mutants. To further investigate the role of the neuronal IIS/FOXO pathway, we identified genes both upregulated by the IIS/FOXO pathway, and genes that are differentially expressed in daf-2 mutants with age. We found that daf-2 differentially-regulated genes in the aged neurons are different from young neurons; in fact, many of these Day 8 daf-2 vs daf-16;daf-2 upregulated genes are stress response and proteolysis genes that may promote neuronal function and health. We then used functional assays to assess the contributions of daf-2-regulated genes to learning and memory. Our results suggest that daf-2’s neuronal targets in older worms are required to maintain neuronal functions with age, suggesting that additional and alternative mechanisms are at work in these aged mutants from their young counterparts.

Results

Wild-type neurons lose their neuronal function and identity with age

Previously, we found that cognitive abilities in C. elegans, including learning, short-term memory, and long-term memory, all decline with age7. Moreover, neuronal morphology and regeneration ability are also impaired with age18,20–22. However, how these phenotypes are regulated at the molecular level in aging neurons remains to be systematically characterized. Therefore, we were interested in first identifying gene expression changes with age in wild-type neurons to characterize the normal physiological aging process. Before choosing timepoints to assess neuronal transcriptome changes, we carried out associative learning and short-term associative memory assays7 as we have previously described7,29–31. Briefly, well-fed worms are starved for one hour, then re-fed while exposed to the neutral odorant butanone for one hour; a choice assay between butanone and control immediately after training tests associative learning, while a choice assay after one hour of recovery on food-only plates tests short-term associative memory7. Adult Day 1 worms are fully developed, young, and healthy, while wild-type Day 7-8 worms, although still in their mid-life, have completely lost their learning and short-term memory abilities already by Day 77 (Fig. 1a), thus we consider them “aged” for the purposes of understanding loss of cognitive ability. Therefore, we reasoned that a comparison of adult wild-type Day 1 neurons with wild-type neurons that are at least aged Day 7-8 should reveal changes with age that result in loss of cognitive function.

Identifying neuronal aging targets in WT worms using neuron-specific RNA-sequencing.

(a) Wild-type learning and 1hr memory results on Day 1 and Day 7. Learning and memory results are represented as learning index (LI). Details of the LI calculation are explained in the methods. Learning, N = 10, memory, N = 5. ****: p < 0.01. Student’s t-test. (b) PCA plot for Day 1 (orange) and Day 8 (blue) neuronal bulk RNA-seq samples. (c) Volcano plot comparing age-associated differentially-expressed genes in WT neurons. Genes downregulated with age (orange) and upregulated with age (blue) were obtained by neuron-specific RNA sequencing of adult wild-type animals with neuron-specific GFP expression. Adjusted p-value < 0.001, log2(Fold-change) > 2. N = 6 biological replicates per age. 1146 genes were significantly downregulated with age (higher in young neurons) and 2016 genes were upregulated with age (higher in old neurons) (d) Tissue prediction scores for genes higher in young neurons. (e) GO terms of genes that decline with age in wild-type neurons. Synaptic and signaling GO terms are enriched in neuronal genes. p-value calculated using hypergeometric distribution probability. (f) Comparison of whole-body higher-in-young genes and neuronal higher-in-young genes. GO Terms and representative genes were performed using g:Profiler software. P-value of overlapping regions were calculated using a hypergeometric calculator. (g) Normalized reads of ins-6, unc-4, mec-7, folt-2, fbf-1, and madd-4, in Day 1 and Day 8 neurons in our dataset. P-adjusted values were calculated from DESeq2 software. Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values.

To identify genes that regulate age-related morphological and functional decline in wild-type neurons, we performed neuron-specific transcriptomic analysis using our previous FACS neuronal isolation method24 on six biological replicates each of Day 1 and Day 8 adult wild-type worms, where 100,000 GFP+ cells were collected for each sample (Fig. S1a-c). Because we previously found that whole-worm analyses mask changes found specifically in neurons24, to complement our aging neuron studies, we also carried out RNA-sequencing analyses of aging whole worms (Fig. S2a-e), which we found is dominated by changes in the extracellular matrix (Fig. S2b), stress response/pathogen genes (Fig. S2c) and the alimentary system (intestine) (Fig. S2d), overshadowing neuronal changes.

Principal components analysis of the FACS-isolated neuron RNA-seq samples indicated that they are well separated by age (Fig. 1b), and downsampling analysis32 (Fig. S1e) suggested that we have sequenced to saturation, with an average of 41,636,463 uniquely-counted reads and detected the expression of 19725 coding and non-coding genes (log10(TPM)>0.5) (Fig. S1d). Enrichment analysis of genes that are differentially expressed with age (Fig. 1c-e) suggested that neuronal sorting and sequencing were successful because the sequenced genes are enriched for neuronal genes such as mec-7, mec-12, and twk-49, as expected, and less enriched for all other major tissues. Tissue enrichment analysis of differentially-expressed genes suggested that aging neurons lose genes most expressed in the nervous system and neurons (Fig. 1d), as one might expect. Gene ontology (GO) analysis suggested that genes declining with age in neurons encode proteins important in neuronal function (Fig. 1e, f), including synaptic proteins (e.g., srh-59, rab-3, sng-1, sup-1), potassium channels (e.g., egl-23, twk-7, twk-49, ncs-5), and transmembrane transporters (e.g., folt-2, ccb-2, unc-79, exp-1). The decrease in expression of these genes during aging may indicate that neurons are losing their identity and their ability to perform neuronal functions, such as signal transduction and axonal transport, and correlates with the behavioral and morphological declines observed in aging wild-type worms.

Comparing whole-worm sequencing and neuron-specific sequencing (Fig. 1f), we found that genes involved in metabolic processes decline with age only in the body, and genes encoding structural proteins, lipid localization, and muscle system processes decline with age in both the body and in neurons, while neurons specifically lose genes that are associated with neuronal function, including synaptic proteins, neuropeptide signaling, and other neuron functions, correlating with neuronal loss of function with age. Together, these results indicate that neurons harbor many unique age-related changes that could be overshadowed in the whole-worm transcriptome but are revealed by neuron-specific sequencing.

Many genes that are more highly expressed in young neurons are known to be specific to a subset of neurons. ins-6, an insulin-like peptide specific to the ASI, ASJ, and AWA neurons28 that regulates longevity33 and aversive learning34, is significantly downregulated with age (Fig. 1g). srd-23, a serpentine receptor located at the AWB neuron cilia35, also decreases expression with age (Fig. S1f). Furthermore, various genes specific to sensory neurons (txt-12, flp-33), touch neurons (mec-7), and motor neurons (unc-4) decline in expression with age (Fig. 1g, Fig. S1f). Previous studies showed that loss of genes including ins-634, mec-736, unc-4, folt-219, madd-437, and fbf-131 lead to behavioral dysfunction in motility and chemosensory abilities; therefore, the decreased expression of these neuron type-specific genes with age may impact the function of individual neurons and disrupt neural circuit communication, ultimately contributing to the declines in behavior observed during aging (Fig. 1g).

As neurons age, genes that increase in expression, while assigned to the nervous system (Fig. 2a) are not specific for neuron function; instead, aged wild-type neurons express higher levels of many predicted F-box genes with predicted proteasome E3 activity (e.g., F-box proteins fbxa-158, fbxb-51, pes-2.1, and SKp1-related proteins skr-12, skr-6). Some transcription regulation (e.g., ced-13, tbx-43, nhr-221, end-1), and chromatin structure and function (e.g., his-54, dot-1.2, jmjd-3.2, hil-7, utx-1) genes also increase with age (Fig. 2b), even though neurons appear to lose their neuron-specific transcriptional identity with age.

Genes that increase with age cause behavioral defects.

(a) Tissue prediction score for wild-type genes expressed at higher levels in aged worms. (b) GO terms of genes expressed higher in aged neurons highlight transcription regulation and proteolysis. GO term analysis was done using Wormcat 2.0. (c) Normalized reads of utx-1 on Day 1 and Day 8. (d) Short-term associative memory (STAM) assay shows that neuron-sensitized adult-only utx-1 knockdown improves 1hr and 2hr memory of wild-type worms on Day 2. RNAi was performed using the neuron-RNAi sensitized strain LC108. (e) Normalized reads of ins-19 on Day 1 and Day 8. (f) ins-19 mutation improves learning and memory in STAM on Day 3 of adulthood. (g) Normalized reads of nmgp-1 on Day 1 and Day 8. (h) nmgp-1 neuron-sensitized RNAi knockdown improves memory in STAM on Day 2. RNAi was performed using the neuron-RNAi sensitized strain LC108.P-adj value of normalized count change generated from DEseq2 analysis. (c,e,g) Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values. Normalized reads and adjusted p value calculated using the DESeq2 software. Each dot represents one sequencing replicate. (d,f,h) N = 5 plates in each behavioral experiment. Representative result of 2 biological repeats is shown. *: p<0.05. **: p<0.01. ***: p<0.001. ****: p<0.0001. Two-way ANOVA with Tukey’s post-hoc analysis.

One ongoing discussion about changes during aging is how to interpret an increase in expression with age. There are two main models for genes that increase their expression with age and have a resulting impact on function: that they rise with age to compensate for lost function (“compensatory”) and therefore promote function, or that their expression is deleterious to function and only rises with age through dysregulation. If a gene is compensatory, then its knockdown would abrogate learning and memory, even in young animals. If a gene’s function is harmful to neurons, reducing its expression might be beneficial to the worm, even in young animals. (Of course, there may be other scenarios in which a gene with multiple functions may be detrimental for some behaviors but beneficial for other physiological roles.) To test this hypothesis, we reduced expression of a small set of highly-upregulated candidate genes in categories that might function in a compensatory manner. These include utx-1, a histone demethylase known to play a role in development38 and lifespan in worms39–41, and whose homolog has been implicated in cognition in mammals42,43; ins-19, an insulin-like peptide; and nmgp-1, a neuronal glycoprotein involved in chemosensation44. In each case, we see that gene expression is significantly higher in old than in young neurons (Fig. 2c, e, g). If a gene increases expression to benefit neurons, we would expect to see no difference in memory in young animals where there is no defect; by contrast, if the increase of a gene is deleterious, we would expect to see an improvement in behavior when knocked down, even in young animals. We performed adult-only neuron-sensitized RNAi knockdown to prevent any possible deleterious effects caused by changes during development, which largely takes place in early larval stages; testing in young adult animals is logical because there is no memory in aged wild-type, so any deleterious effect of knocking down a potentially compensatory gene in an aged would not result in a change. For all behavioral assays, we first prioritized significantly-changed genes with high fold-change, and then those with mammalian homologs.

We found that 48 hrs (L4-Day 2) of adult-only knockdown of utx-1 increases 1hr and 2hr memory (Fig. 2d), the loss-of-function mutation of ins-19 increases both learning and memory (Fig. 2f) and the adult-only knock-down of nmgp-1 extends memory at 2 hours (Fig. 2h). That is, in each of these cases, reduction of these genes did not impair memory, as loss of a compensatory function would appear; rather, loss of these age-upregulated genes improved wild-type memory. These results indicate that at least some neuronal genes that increase with age can have a negative impact on learning and memory, as demonstrated by the improvement of memory when knocked down, even in young animals. While it is still possible that some upregulated genes may act in a compensatory manner, the simplest model is that at least some are actively deleterious for learning and memory. We previously observed that for genes that play a role in complex behaviors like learning and memory, the loss of single genes can have a large impact on these complex behaviors60, unlike the additive roles of longevity-promoting genes 11. Therefore, one mechanism by which wild-type worms lose their learning and memory functions with age is not just by loss of neuronal gene expression, as one might expect, but also by dysregulation of expression of genes that can negatively impact learning and memory.

daf-2 mutants maintain learning and memory with age

We previously found that daf-2 animals have extended motility (Maximum Velocity) that correlates with and predicts their extension of lifespan8. Additionally, not only do young daf-2 worms have better memory than wild-type worms, but daf-2 mutants also maintain learning and memory better with age7,24. However, the duration of this improvement was unknown. To determine the proportion of life that worms can learn and remember, we tested wild-type, daf-2, and daf-16;daf-2 worms for their learning and associative memory ability every day until these functions were lost. We found that while wild-type worms lose their learning and short-term memory abilities by Day 7-8 (Fig. 1a, Fig. 3a, b), learning and memory span were significantly extended in daf-2 mutants (Fig. 3a, b); thus, a comparison of daf-2 neurons with wild-type neurons at Day 8 should reveal differences relevant to cognitive aging. The extension of learning and memory is dependent on the FOXO transcription factor DAF-16 (Fig. 3a); in fact, while daf-16;daf-2 mutants still have the ability to learn for a few days, these mutants are completely unable to carry out any memory ability, even on Day 1. Thus, learning ability, which is similar in wild-type and daf-2;daf-16 mutants, is mechanistically distinct from short-term memory ability31. daf-2 worms maintained learning ability until Day 19 and short-term (1hr) memory ability until Day 15, more than twice the duration of wild-type worms, while daf-16;daf-2 worms exhibit no short-term memory ability, even on Day 1 of adulthood (Fig. 3b). Our data suggest that the learning span-to-lifespan (Fig. S3c) and memory span-to-lifespan ratios in daf-2 worms were similar to or slightly higher than that of wild-type worms (Fig. 3c, d), indicating that daf-2 mutants maintain cognitive function for at least proportionally as long as wild-type worms do. Thus, daf-2 mutants maintain their higher cognitive quality of life longer than wild-type worms, while daf-16;daf-2 mutants spend their whole lives without memory ability (Fig. 3d), in contrast to claims that daf-2 mutants are less healthy than wild-type or daf-16 worms23. Additionally it should be noted that because our choice assays distinguish motility function from learning and memory function29, the improvements in memory with age shown by daf-2 mutants relative to wild type are distinct from daf-2’s improvements in motility that we previously showed8. Therefore, we are interested in these genes that might contribute to the extended cognitive function that daf-2 worms demonstrate.

Identifying neuronal IIS/FOXO targets in aged worms using neuron-specific RNA-sequencing.

(a) daf-2 mutants show better learning maintenance with age compared to N2 and daf-16;daf-2 worms. N = 10 plates in each condition. (b) daf-2 mutants show better memory maintenance with age compared to N2 worms. daf-16;daf-2 worms do not have 1hr memory on Day 1 of adulthood. N = 10 plates in each condition. (c-d) daf-2 mutants have a slightly larger learning span/lifespan ratio and memory span/lifespan ratio than N2 (wild type). Lifespan shown in Figure S3c. (e) PCA plot of Day 8 N2, daf-2, and daf-16;daf-2 neuronal RNA sequencing results. (f) Volcano plot of neuronal daf-2-regulated, daf-16-dependent up- and downregulated genes on adult Day 8 (Adjusted p-value < 0.05, log2(Fold-change) > 0.5, N = 6 biological replicates per strain). 570 genes were significantly upregulated and 814 genes were downregulated in daf-2 neurons compared with daf-16;daf-2. (g) Volcano plot of whole-worm daf-2 vs daf-16;daf-2 differentially-expressed genes during aging. 3154 genes are higher in daf-2, 1289 genes are higher in daf-16;daf-2 (log2[Fold-change(daf-2 vs daf-16;daf-2)] >1.5, p-adjusted <0.01). (h) Comparison of neuronal and whole-worm Day 8 daf-2 differentially-expressed genes (overlap p = 3.34E-63, hypergeometric test). Neuron-specific and shared daf-2 upregulated genes with the highest fold-changes are labeled.

Aging IIS/FOXO neurons express stress-resistance genes to maintain neuronal function with age

To identify genes that may improve memory and slow cognitive aging in long-lived daf-2 mutants, we compared the transcriptional profiles of Day 8 FACS-isolated neurons from daf-2 animals with Day 8 FACS-isolated wild-type and daf-16;daf-2 neurons; by Day 8, wild-type and daf-16;daf-2 worms have already lost their learning and memory ability, but daf-2 worms still maintain their cognitive functions (Fig. 3a, b). It should be noted that wild-type worms still have normal chemotaxis and motility at Day 87, and there is a separation of several days between the loss of cognitive functions and the loss of motility7; therefore, comparison of the neuronal transcriptomes of daf-2 with wild type and daf-16;daf-2 at this age should specifically highlight genes that are required for learning and memory rather than other functions.

The PCA of the daf-2, daf-16;daf-2, and wild-type neuronal Day 8 transcriptomes (Fig. 3e) indicates that aged daf-16;daf-2 mutant neurons are similar to aged wild-type neurons, correlating well with their similarly worsened cognitive functions at this age; that is, at a transcriptomic level, aged (Day 8) wild-type neurons and aged daf-16;daf-2 neurons are similar, which is echoed by their shared inability to carry out learning and memory by Day 8 of adulthood (Fig. 3ea, b). By contrast, the transcriptomes of aged daf-2 mutant neurons are distinct from both aged wild-type and aged daf-16;daf-2 neuron transcriptomes, just as the cognitive abilities of daf-2 are much greater than wild type or daf-16;daf-2 at this age. Downsampling analysis shows that our sequencing depth is sufficient to saturate the detectable differential expression (Fig. S3g, h). We obtained an average of 47,233,119 counted reads per sample (Supplemental Table 5) and detected the expression of 16,488 coding and non-coding genes (Fig. S3e).

We identified 570 upregulated and 814 downregulated genes in Day 8 daf-2 neurons compared to Day 8 daf-16;daf-2 neurons (Fig. 3f). A large fraction of the downregulated genes in Day 8 daf-2 vs daf-16;daf-2 neurons are “nematode-specific peptide family” (nspc-) genes of unknown function (Fig. 3f). While the daf-2 vs daf-16;daf-2 changes in whole worms largely replicated the results from our previous studies of young animals11,45 (Fig. 3g, Fig S6g), comparison of the daf-2 vs daf-16;daf-2 differential transcriptional changes in Day 8 whole worms and Day 8 neurons reveal shared (174 genes) and neuron-specific gene expression changes (396 genes; dod-24, srh-2, lin-42, etc.) (Fig. 3h, Fig. S4c). Not surprisingly, previously-identified genes from whole-worm daf-2 vs daf-16;daf-2 and N2 (e.g., sod-3, mtl-1, cpi-1, hsp-12.6, etc.) that play roles in both neurons and other tissues even in Day 1 daf-2 mutants appear in the shared list. (Some neuron-specific Day 8 daf-2-upregulated genes have not been reported to be expressed in neurons previously (e.g., spin-2), further suggesting the value of transcriptomic analyses of isolated neurons in mutant backgrounds at this age.)

Many genes upregulated in Day 8 daf-2 neurons relative to daf-16;daf-2 are related to stress responses, including heat stress (e.g., hsp-12.6, hsp-12.3, F08H9.4/hsp), oxidative stress (e.g., sod-3), and metal stress genes (e.g., mtl-1); and proteolysis (e.g., cpi-1, cpr-2, and tep-1). The upregulation of these genes may perform neuroprotective functions, as their homologs in mammals have been shown to do (Table 1). Specifically, 36 of the top 100 upregulated genes have identified orthologs or identified domains with known function, of which 32 of (89%) have functions in promoting neuronal health. These mammalian homologs protect neurons against protein aggregation and harmful metabolites (e.g., cpi-1, alh-2, ttr-41, gpx-5)46–50, maintain synaptic organization and neuronal homeostasis (e.g., dod-24, ptr-19, plep-1)51–53, facilitate neuronal injury repair (e.g., F08H9.4, sod-3)54,55, and maintain normal neuronal function (e.g., lgc-28, slc-36.3, lin-42)56–59. Together, these genes may help maintain daf-2’s neuronal health and protect neurons from accumulation of environmental harm during aging.

We found that about a third of the daf-2-upregulated genes were shared between the daf-2 vs daf-16;daf-2 analysis and the daf-2 vs N2 analysis (338 genes) (Fig. S5). Of the unshared genes, the daf-2-maintained genes that are specific to the daf-2 vs N2 comparison are bZIP transcription factors, including zip-5, zip-4, atf-2, and proteasome components (Figure S6D). These results indicate that other transcription factors may participate in regulating daf-2 functions in aged neurons in addition to the daf-16/FOXO transcription factor.

IIS/FOXO transcriptomic changes are necessary for daf-2 mutant’s improved neuronal functions

We were interested not only in the genes that remained upregulated with age, but also in genes that might have increased with age in the high-performing daf-2 mutants. That is, are there genes that increase in expression in daf-2 mutants that are necessary or beneficial for their continued high performance with age? Some of the Day 8 daf-2 vs wild-type or daf-16;daf-2 upregulated genes are also Class 1 DAF-16-dependent genes11 (sod-3, hsp-12.3, fat-5, and mtl-1, hil-1, and dao-2). However, many more genes were differentially expressed in Day 8 daf-2 vs daf-16;daf-2 neurons from our Day 1 data24 (Fig. 4a, Fig. S4d). Of the “new” genes – that is, genes upregulated specifically in neurons of Day 8 vs Day 1 of daf-2 vs daf-16;daf-2 – many have mammalian homologs that have been shown to play neuroprotective roles, by protecting against aggregation proteins and harmful metabolites, maintaining synaptic organization, neuronal homoeostasis, or neuronal activity, or facilitating neuronal injury repair (see Table 1 for specific references).

Neuronal IIS/FOXO aging targets regulate memory decline with age in daf-2 worms. (a) Comparison of neuronal Day 1 and Day 8 daf-2 vs daf-16;daf-2 upregulated genes. All shared genes and top Day 8-specific daf-2 upregulated genes are labeled. (b) daf-2-regulated fold-change profile of candidate genes. All candidates are upregulated in daf-2 mutants. (c) Description of candidate genes. log2(Fold-change) and p-adjusted values from the daf-2 vs daf-16;daf-2 comparison unless stated otherwise. (d) Candidate gene knockdown effects on Day 6 adult daf-2 learning (0hr after conditioning). Two candidate genes, dod-24 and F08H9.4, show a significant decrease in learning ability. N = 5 plates in each condition, merged results of 3 biological repeats shown. (e) Candidate gene knockdown effects on Day 6 adult daf-2 short-term memory (1hr after conditioning). C44B7.5, dod-24, F08H9.4, mtl-1, and alh-2 showed significant decreases in memory. N = 5 plates in each condition, the representative image of 3 biological repeats shown. (d-e) RNAi was performed using a neuron-sensitized RNAi strain CQ745: daf-2(e1370) III; vIs69 [pCFJ90(Pmyo-2::mCherry + Punc-119::sid-1)] V .*: p < 0.05. **: p < 0.01. ***: p<0.001. ****: p <0.0001. One-way ANOVA with Dunnet’s post-hoc analysis. Box plots: center line, median; box range, 25-75th percentiles; whiskers denote minimum-maximum values.

If the upregulated genes in aged daf-2 neurons are responsible for the extended memory span of daf-2 mutants, knocking down those genes should block older daf-2 mutants’ memory functions. Therefore, we tested the effect of RNAi knockdown of the top fold-change candidate genes on daf-2’s memory in aged adults. We chose Day 6 for testing because by then, like on Day 8, wild-type worms have already lost their learning and most memory abilities, but daf-2 worms retain normal cognitive functions, and this time point avoids the increased naïve chemotaxis that we observe in older daf-2 animals. As shown in Fig. 4b-c, we selected these significantly differentially-expressed candidate genes based on their ranking in fold-change. Previously, we have found that the top significantly differentially-expressed genes (by fold-change) are most likely to have strong effects on function, while less differentially-changed genes have less of an effect11,24,60, therefore we prioritized genes that are significantly different and the most highly expressed in daf-2 mutants compared to daf-16;daf-2 mutants for subsequent testing (Fig. 4c). daf-2 worms, including neurons, are susceptible to RNA interference24,61. Of the eight candidate genes we tested, the reduction of three of them (F08H9.4, mtl-1, and dod-24, originally classified as a Class II gene with proposed immune activity) significantly decreased daf-2’s learning ability on Day 6 (Fig. 4d). Those genes plus reduction of two additional genes (C44B7.5 and alh-2) affected 1hr memory (Fig. 4e) in Day 6 daf-2 mutants. That is, knockdown of the heat shock-related gene F08H9.4, the innate immunity gene dod-24, aldehyde dehydrogenase alh-2 and previously uncharacterized gene C44B7.5 are required to some degree for daf-2’s extended memory ability. The reduction of the metal stress gene mtl-1, which is expressed in neurons as well as the rest of the body, had a slight effect on learning and memory.

One caveat of these experiments is that, while we found these genes through the isolation of neurons from aged worms and subsequent RNA-sequencing, the knockdown of the genes and its effects are not necessarily neuron-autonomous; however, alh-2 and F08H9.4 were reported to only be expressed in neurons and the cephalic sheath cell62, and C44B7.5 and dod-24, while expressed more broadly, were not upregulated in daf-2 in the whole-worm analysis (Fig. 3f), therefore their effects are most likely neuron-autonomous. In fact, dod-24 is one of the original Class 2 daf-2-downregulated genes from whole-animal analyses, suggesting that dod-24’s increase in expression is specifically in neurons, therefore the effect of its knockdown is most likely to be neuron-autonomous.

Together, these data suggest that the specific genes that are differentially regulated in Day 8 daf-2 mutants may aid in slowing neuronal function decline and behavioral changes associated with aging. Furthermore, memory maintenance with age might require additional genes that function in promoting stress resistance and neuronal resilience, which were not previously uncovered in analyses of young animals.

Discussion

Although it has been shown previously that daf-2 worms maintain various functions with age, how long they can maintain learning and memory with age, and the genes that might be responsible for these extended neuronal functions, have not been previously explored. Here we have found that daf-2 worms maintain learning and memory abilities proportional with (or even slightly beyond) their degree of lifespan extension, underscoring daf-2’s improved healthspan8. To understand how memory is lost with age and retained in insulin/IGF-1-like signaling mutants, we have characterized the neuronal transcriptomes of aged wild-type worms and IIS (daf-2) and IIS/FOXO (daf-16;daf-2) mutants (Fig. 5). We found that wild-type neuronal aging is characterized by a downregulation of neuronal function genes and an upregulation of proteolysis genes and transcriptional and epigenetic regulators, which together may help explain the loss of neuronal identity and function with age. We also identified the transcriptomic profile accompanying daf-2’s extended learning and memory span. Specifically, daf-2 neurons maintain higher expression of stress response genes and predicted neuronal homeostasis functions (Table 1), which may help make them more resistant to environmental adversities and age-related decline. We also identified genes responsible for wild-type worms’ worsened learning and memory with age.

Aged daf-2 neurons upregulate neuroprotective genes to maintain improved cognitive behaviors.

During normal neuronal aging, neuron-specific genes decrease in expression, while proteolysis and epigenetic regulators are upregulated, resulting in neuron dysfunction and cognitive function loss. In aged daf-2 mutants, upregulation of neuroprotective genes including dod-24, F08H9.4, C44B7.5, alh-2, and mtl-1 contribute to daf-2’s improved cognitive function. The diagram was generated using Biorender.

By employing a FACS-based neuron-sorting technique, we selectively analyzed adult neuron-function-related genes and investigated their aging process, which is not easily discernible through whole-worm sequencing25–27 (Fig. S6). Sequencing many biological repeats of aging neurons to high depth with ribosomal RNA depletion allowed us to detect a larger number of genes compared to other neuron-related bulk and single-cell sequencing profiles26, providing a deep transcriptomic dataset of aged wild-type, IIS mutant, and IIS/FOXO mutant neurons. Our analysis allowed us to identify differentially-expressed genes that are known to be expressed in at a small number of neurons, even for low-abundance genes. Notably, our sequencing results uncovered genes previously not known to be expressed in neurons that remained undetected in other datasets. Moreover, we revealed the involvement of known neuronal genes in the aging process, such as ins-6 and srd-23. We hope that this dataset will become a valuable resource for detecting new candidates in neuronal aging.

For example, dod-24, which we observed to be upregulated in daf-2 neurons and required for daf-2’s extended memory, was downregulated in the daf-2 whole-worm transcriptome (Fig. S4c). dod-24 has been traditionally classified as a Class II gene that is downregulated in daf-2 worms and upregulated by daf-16 RNAi treatment11,45. Functionally, it has been shown to be an innate immunity gene upregulated during pathogen infection63–65, and its whole-body reduction has been shown to extend the lifespan of wild-type animals11. However, here we find that dod-24 is beneficial in the nervous system and required for daf-2’s extended learning and memory in aged worms. This intriguing contrast between the whole-worm transcriptome and the neuron-specific transcriptome suggests that some genes may have distinct regulatory roles in the nervous system, necessitating a more precise approach beyond whole-worm transcriptomics.

Using this neuron-specific sequencing profile of aged cells, we identified key pathways that change during neuron aging. Our sequencing of aged neurons uncovered active transcriptomic alterations during aging, resulting in not just transcriptional silencing but also upregulation of various pathways. We found that the Day 8 daf-2 vs daf-16;daf-2 neuronal differentially-expressed genes that we have newly discovered here differ from the neuronal Day 1 daf-2 vs daf-16;daf-2 dataset we previously obtained24. These Day 8 differentially-expressed genes are not canonical neuronal genes, such as receptors and ion channels; instead, there are more metabolic and proteolytic genes whose protein orthologs have been shown to be neuroprotective. All top 50 genes and 90% of the top 100 genes with identified mammalian orthologs have been shown to be essential to neuronal functions in mammals (Table 1). These results indicate that instead of simply mimicking a young state, daf-2 may enhance neuron’s resilience to accumulation of harm and take a more active approach to combat aging. These changes suggest that daf-2’s extended memory maintenance may require different mechanisms than function in young animals; daf-2 may maintain neuronal function not just by retaining a youthful transcriptome, but also by increasing the expression of genes that promote resilience, such as stress-response genes and proteolysis inhibitors.

In addition to examining aging in wild-type and IIS/FOXO mutants independently, our results further linked the normal aging process to altered gene regulation in the IIS pathway. utx-1, nmgp-1, and ins-19 increase in expression in aged neurons, and we found that their reduction improved memory, indicating that at least some of the genes whose expression rises with age can have a negative impact on normal cognitive functions, rather than acting in a compensatory manner. utx-1 is an H3K27me3 histone demethylase we found to be higher in wild-type aged neurons, but it is also involved in the IIS pathway. The downregulation of utx-1 has been shown to regulate development38 and promote longevity39–41, and its mammalian homolog has been implicated in regulating cognitive abilities42,43. The longevity response of utx-1 depends on daf-1639–41. The loss of utx-1 decreases methylation on the daf-2 gene, thus increasing DAF-16’s nuclear localization, mimicking a daf-2 mutation39. This example of the crosstalk between normal aging and IIS/FOXO mutants offers valuable insights into modifying the aging process for enhanced longevity and cognitive health.

We found that the insulin-like peptide ins-19 was upregulated in aged neurons and was downregulated in aged daf-2 neurons, and its downregulation in wild-type worms extended memory span. Insulin-like peptides play crucial roles as receptor ligands (in both agonist and antagonist roles) for DAF-2, and we have found them to be downregulated in daf-2 mutants compared with daf-16;daf-2 mutants, possibly creating a feedback loop that dampens the insulin signaling pathway, as was previously shown for ins-7 and ins-1811,66. These peptides exhibit diverse functions in development, dauer formation, and longevity11,66–70. Notably, certain insulin-like peptides have been linked to neuronal activities, such as the regulation of aversive learning by the two antagonistic peptides ins-6 and ins-734, and reduced long-term learning and memory by ins-22 RNAi60. In our study, the expression changes of ins-19 during wild-type aging and in daf-2 mutants provide an example of how longevity mutants can reverse wild-type transcriptional changes during aging, ultimately reducing behavioral and functional decline.

Summary of mechanistic insights

Our analysis of transcriptomes from isolated aged Day 8 neurons of wild-type, daf-2, and daf-16;daf-16 mutants reveal several major mechanistic insights. Specifically, we found that wild-type neurons lose their neuronal identity through a combination of the loss of neuron-specific function genes with age, and the concomitant dysregulated increase in non-neuronal genes with age. Furthermore, at least a fraction of the top-upregulated genes with age can play deleterious roles; that is, they rise with age, and their knockdown improves function, even in young animals. This argues against the idea that all of these genes are playing a compensatory role with age.

We also found that the knockdown of individual top-ranked genes that function in learning and memory can have a large impact -like removing a cog of a machine. This is in contrast to our earlier findings regarding gene reduction in lifespan, where most cellular longevity processes regulated by DAF-16 activity appear to be additive, and therefore loss of individual major genes downstream of DAF-2 and DAF-16 have at most a 5-10% impact11. Several of these genes we found to be required for daf-2’s age-related improvement in learning and memory - namely dod-24, F08H9.4, C44B7.5, and alh-2 – were previously not associated with memory function. Finally, these genes are distinct from the set of upregulated Day 1 daf-2 vs daf-16;daf-2 genes; how they each individually maintain neuronal function better with age will be interesting to dissect.

Conclusions

Beyond our sequencing analysis, we have established links between genomics, function, and behavior. We also identified several new genes required for daf-2’s age-related improvement in learning and memory, shedding light on their neuron-specific roles. These additional findings further suggest that neuronal sequencing datasets can be used to identify functional candidates and pathways during the aging process. By bridging the gap between transcriptomic landscapes, genetic regulation, and functional outcomes, our study provides a greater understanding of the mechanisms underlying neuronal aging, providing insights into development of aging interventions.

Methods

Strains and worm cultivation

N2 (wild type), OH441: otIs45(unc-119::GFP), CQ295: otIs45(unc-119::GFP);daf-2(e1370), CQ296: otIs45(unc-119::GFP) ;daf-16(mu86);daf-2(e1370), LC108: uIs69 (myo-2p::mCherry + unc-119p::sid-1), CQ705: daf-2(e1370) III, 3X outcrossed, CQ745: daf-2(e1370) III; vIs69 [pCFJ90(Pmyo-2::mCherry + Punc-119::sid-1)] V, QL188: ins-19(tm5155) II, CX3695: kyIs140(str-2::GFP + lin-15(+)), CQ461: (daf-2(e1370);Pmec-4::mCherry) and CQ501: (daf-2 (e1370);zip-5(gk646);Pmec-4::mCherry). Strains were grown on high-growth media (HGM) plates seeded with E. coli OP50 bacteria using standard methods 71.

Tissue-specific isolation

For neuronal isolation, five plates of fully-grown worms from HG plates were synchronized by hypochlorite treatment, eggs spread on seeded plates to hatch, and at least five plates/replicate were grown to L4 on HGM plates until transferred to HGM plates with FUdR to avoid progeny contamination. This gives us ∼6000 healthy Day 8 worms to sort. Neuron isolation and Fluorescent-activated cell sorting were carried out as previously described 24,62. Briefly, worms were treated with 1000uL lysis buffer (200 mM DTT, 0.25% SDS, 20 mM HEPES pH 8.0, 3% sucrose) for 6.5 mins to break the cuticle. Then worms were washed and resuspended in 500uL 20mg/mL pronase from Streptomyces griseus (Sigma-Aldrich). Worms were incubated at room temperature with mechanical disruption by pipetting until no whole-worm bodies were seen, and then ice-cold osmolarity adjusted L-15 buffer(Gibco) with 2% Fetal Bovine Serum (Gibco) were added to stop the reaction. Prior to sorting, cell suspensions were filtered using a 5um filter and sorted using a FACSVantage SE w/ DiVa (BD Biosciences; 488nm excitation, 530/30nm bandpass filter for GFP detection). Sorting gates were determined by comparing with age-matched, genotype-matched non-fluorescent cell suspension samples. Fluorescent neuron cells were directly sorted into Trizol LS. 100,000 GFP+ cells were collected for each sample.

RNA extraction, library generation and sequencing

We used standard trizol-chloroform-isopropanol method to extract RNA, then performed RNA cleanup using RNeasy MinElute Cleanup Kit (Qiagen). RNA quality was assessed using the Agilent Bioanalyzer RNA Pico chip, and bioanalyzer RIN>6.0 samples were observed before library generation. 2ng of RNA was used for library generation using Ovation SoLo RNA-Seq library preparation kit with AnyDeplete Probe Mix-C. elegans (Tecan Genomics) according to the manufacturer’s instructions 72. Library quality and concentration was assessed using Agilent Bioanalyzer DNA 12000 chip. Samples were multiplexed and sequencing were performed using NovaSeq S1 100nt Flowcell v1.5 (Illumina).

Data processing

FastQC was performed on each sample for quality control analysis. RNA STAR package was used for mapping paired-end reads to the C. elegans genome ce11 (UCSC Feb 2013) using the gene model ws245genes.gtf. Length of the genomic sequence around annotated junctions is chosen as read length -1. 50-70% reads were uniquely mapped. Reads uniquely mapped to the genome were then counted using htseq-count (mode = union). DESeq2 analysis was then used for read normalization and differential expression analysis on counted reads 73. Genes with a log10TPM >0.5 were considered as detected and genes with a log2(fold-change) > 0.5 and p-adjusted <0.05 are considered differentially expressed in further analysis. Gene ontology analysis were performed using gprofiler74 or WormCat 2.075 and category 2 was selected to show. Tissue query was performed on top 500 highest fold-change genes, using worm tissue query website (www.worm.princeton.edu)62, and only major systems were selected in the analysis.

Learning and Memory Experiments

We performed Short-Term Associative Memory (STAM) experiments as previously described 7,29. Briefly, we used five plates of synchronized adult worms/sample to perform each experiment. One plate of worms was used to test naïve chemotaxis assay without conditioning, while the other 3 plates were washed into M9, and washed 3 additional times to get rid of the bacteria. These worms are starved for 1hr to prime them for food uptake. Then these worms are transferred to conditioning plates with NGM plates seeded with OP50 and 10% butanone stripes on the lid for 1hr to perform conditioning. After conditioning, worms are either transferred from the conditioning plate directly to the chemotaxis plates to assess learning, or transferred to holding plate for 1hr or 2hrs to assess for memory. After staying on holding plates for 1hr or 2hrs, worms are then transferred onto chemotaxis plates to assess for short-term memory. Chemotaxis assays were performed by transferring worms onto chemotaxis plates with 1uL 10% butanone and 1uL ethanol spots separated by 8cm on a 10cm NGM plate. Worms who have reached either butanone spot or the ethanol spot are paralyzed by the 1uL 7.5% NaN3 on these spots. For each timepoint, 5 chemotaxis plates are used to minimize the variation of the outcome. We performed this chemotaxis assay to butanone on naïve and appetitive-trained worms at different time points to assess change in preference to butanone. Chemotaxis index is calculated as (# of worms at butanone-# of worms at ethanol)/(total # of worms - # of worms at origin).

Learning index is calculated by subtracting trained chemotaxis index with naïve chemotaxis index.

For learning and memory span assays, we obtained synchronized worms from hypochlorite-treated eggs. Synchronized worms were washed onto 5’-fluorodeoxyuridine (FUdR) at L4 and maintained on FUdR plates by transferring to new plates every 2 days. 1 Day Prior to experiments, worms are washed onto fresh HG plates without FUdR to avoid change in behavior caused by FUdR. To verify that FUdR has no effect on short-term memory, we compared worms with and without FUdR (Fig. S6d), and found no differences. For utx-1 and nmgp-1 RNAi experiments, synchronized L4 neuron-RNAi sensitized worms were washed onto HGM plates with carbenicillin and IPTG and seeded with HT115 RNAi bacteria containing the RNAi constructs from the Ahringer Library. For daf-2 upregulated candidates’ RNAi experiments, synchronized L4 daf-2 neuron-RNAi-sensitized worms were washed onto HGM plates added with carbenicillin, FUdR, and isopropyl-b-D-thiogalactopyranoside (IPTG) and seeded with HT115 bacteria containing RNAi constructs generated from the Ahringer RNAi Library, then were transferred onto fresh RNAi plates every 2 days until Day 6. 1 Day Prior to experiments, worms are transferred onto plates without FUdR.

Quantitative and Statistical Analysis

All experimental analysis was performed using Prism 8 software. Two-way ANOVA with Tukey post-hoc tests were used to compare the learning curve between control and experimental groups. One-way ANOVA followed by Dunnet post-hoc tests for multiple comparisons was performed to compare learning or 2hr memory between various treatment groups and control. Chi-square test was performed to compare the neuron morphology change between young and aged AWC neurons. All GO term analysis were perform using Wormcat 2.0 software with Bonferroni corrected adjusted p-values. Venn diagram overlaps were compared using the hypergeometric test. Differential expression analysis of RNA-seq were performed using DESeq2 algorithm and adjusted p-values were generated with Wald test using Benjamini and Hochberg method (BH-adjusted p values). Additional statistical details of experiments, including sample size (with n representing the number of chemotaxis assays performed for behavior, RNA collections for RNA-seq, and the number of worms for microscopy), can be found in the methods and figure legends. Regression analyses were performed using sklearn packages. Correlations were calculated using the SciPy packages.